---
title: "analysis_supp_fig6"
author: '500455892'
date: '`r Sys.Date()`'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Analysis for supp fig 6: comparison vs spatial DR methods on PRECAST's paper's simulated data

Packages and data

Note: you will need to download the data yourself to run this analysis. See the data availability section of the manuscript. 
```{r}
library(anndata)
library(tidyverse)
library(PRECAST)
library(SeuratObject)
library(wSIR)
library(DR.SC)
library(vctrs) # for vec_rep_each function
library(SpatialPCA)
library(doBy)
library(nnet)
library(DescTools)
library(scran)
library(scuttle) 
library(scater)
library(patchwork)
library(vctrs)
source("spatial_benchmark_functions.R")
dlpfc_spe <- readRDS("dski/nobackup/biostat/datasets/SpatialTranscriptomics/Visium_DLPFC/spe_DLPFC.Rds")
```

STAMP analysis

set up python
```{r}
use_python("/Library/Frameworks/Python.framework/Versions/3.8/bin/python3") # run this one
test_file <- "stamp_wrapper.py"
source_python(test_file)

stamp_wrapper_complete <- function(X, coords) {
  out <- stamp_wrapper_fn(counts = as.matrix(X), 
                          coordinates = coords)
  stamp_cells_taken <- as.numeric(out[[2]]) + 1 # +1 since python 0-indexed
  stamp_coords_taken <- coords[stamp_cells_taken,]
  return(list(embedding = out[[1]],
              coords = stamp_coords_taken))
}
```

run STAMP analysis

```{r}
samples_vec <- unique(spe_DLPFC@colData$sample)
psR2_stamp_vals <- c()
 
for (current_sample in samples_vec) {
  # extract this sample
  spe_DLPFC_currentsample <- spe_DLPFC[,spe_DLPFC@colData$sample == current_sample]
  
  # filter HVGs
  modelled_var_v1 <- modelGeneVar(spe_DLPFC_currentsample@assays@data$logcounts)
  topgenes_v1 <- scran::getTopHVGs(modelled_var_v1, n = 500)
  spe_DLPFC_currentsample <- spe_DLPFC_currentsample[topgenes_v1,]
  
  # extract current coords and counts
  coords1 <- spe_DLPFC_currentsample@int_colData@listData$spatialCoords
  counts1 <- t(spe_DLPFC_currentsample@assays@data$counts)
  domains1 <- spe_DLPFC_currentsample@colData$layer_guess_reordered_short

  stamp_out <- stamp_wrapper_complete(X = counts1,
                                      coords = as.matrix(coords1))
  stamp_cells <- domains1[(spe_DLPFC_currentsample@colData$spot_ID %in% rownames(stamp_out$coords))]
  fit_stamp <- multinom(stamp_cells ~ as.matrix(stamp_out$embedding), maxit = 1000, MaxNWts = 2000,model = TRUE, verbose = FALSE)

  psR2_stamp_vals <- psR2_stamp_vals %>% append(PseudoR2(fit_stamp, c("McFaddenAdj")))
}
```

Other method analysis

```{r}
samples_vec <- unique(spe_DLPFC@colData$sample)

psR2_wsir_vals <- c()
psR2_precast_vals <- c()
psR2_drsc_vals <- c()
psR2_spca_vals <- c()
 
for (current_sample in samples_vec) {
  # extract this sample
  spe_DLPFC_currentsample <- spe_DLPFC[,spe_DLPFC@colData$sample == current_sample]
  
  # filter HVGs
  modelled_var_v1 <- modelGeneVar(spe_DLPFC_currentsample@assays@data$logcounts)
  topgenes_v1 <- scran::getTopHVGs(modelled_var_v1, n = 500)
  spe_DLPFC_currentsample <- spe_DLPFC_currentsample[topgenes_v1,]
  
  # extract current exprs, coords, counts
  exprs1 <- t(spe_DLPFC_currentsample@assays@data$logcounts)
  coords1 <- spe_DLPFC_currentsample@int_colData@listData$spatialCoords
  counts1 <- t(spe_DLPFC_currentsample@assays@data$counts)
  domains1 <- spe_DLPFC_currentsample@colData$layer_guess_reordered_short
  
  current_k <- length(unique(spe_DLPFC_currentsample@colData$layer_guess_reordered_short))

  # create low dims
  lowdim_out <- lowdims_resub(exprs = exprs1,
                              coords = coords1,
                              K = current_k)
  spca_out <- spca_wrapper(counts = t(counts1), coords = coords1)
  
  # extract spatial domains for precast and spca
  precast_domains <- domains1[(spe_DLPFC_currentsample@colData$spot_ID %in% rownames(lowdim_out$PRECAST))]
  spca_domains <- domains1[(spe_DLPFC_currentsample@colData$spot_ID %in% rownames(spca_out$coords))]
  
  # fit multilogistic regression models on each low dim
  fitwsir <- multinom(domains1 ~ lowdim_out$wSIR, maxit = 1000, MaxNWts = 2000,model = TRUE, verbose = FALSE)
  fitprecast <- multinom(precast_domains ~ lowdim_out$PRECAST, maxit = 1000, MaxNWts = 2000,model = TRUE, verbose = FALSE)
  fitdrsc <- multinom(domains1 ~ lowdim_out$DRSC, maxit = 1000, MaxNWts = 2000,model = TRUE, verbose = FALSE)
  fitspca <- multinom(spca_domains ~ spca_out$projected, maxit = 1000, MaxNWts = 2000,model = TRUE, verbose = FALSE)
  
  # save out the McF-psR2 value from each method
  psR2_wsir_vals <-  psR2_wsir_vals %>% append(PseudoR2(fitwsir, c("McFaddenAdj")))
  psR2_precast_vals <- psR2_precast_vals %>% append(PseudoR2(fitprecast, c("McFaddenAdj")))
  psR2_drsc_vals <- psR2_drsc_vals %>% append(PseudoR2(fitdrsc, c("McFaddenAdj")))
  psR2_spca_vals <- psR2_spca_vals %>% append(PseudoR2(fitspca, c("McFaddenAdj")))
}

#load("DLPFC_results/psR2_stamp_vals.RData")
psR2_dlpfc_df_full <- data.frame(psR2 = c(psR2_wsir_vals,
                                          psR2_precast_vals,
                                          psR2_drsc_vals,
                                          psR2_spca_vals,
                                          psR2_stamp_vals),
                                  sample = rep(samples_vec, 5),
                                  method = vec_rep_each(c("wSIR", "PRECAST", "DR.SC", "SpatialPCA", "STAMP"), nsamp))
save(psR2_dlpfc_df_full, file = "psR2_dlpfc_df_full.RData")
```

# ARI

create wrapper functions for clustering

```{r}
library(aricode)
precast_wrapper_cluster <- function(exprs, 
                                    coords, 
                                    K = 7, 
                                    maxIter = 30, 
                                    gene.number = ncol(exprs)) {
  seurat_obj <- CreateSeuratObject(counts = t(exprs), project = "SpatialProject")
  row.names(coords) <- rownames(exprs)
  colnames(coords) <- c("row", "col")
  seurat_obj <- AddMetaData(seurat_obj, metadata = coords)
  
  preobj <- CreatePRECASTObject(seuList = list(seurat_obj), 
                                selectGenesMethod = "HVGs", 
                                gene.number = gene.number,
                                verbose = FALSE) 
  PRECASTObj <- AddAdjList(preobj, platform = "ST")
  PRECASTObj <- AddParSetting(PRECASTObj, 
                              Sigma_equal = FALSE, 
                              coreNum = 1, 
                              maxIter = maxIter, 
                              verbose = FALSE)
  
  PRECASTObj <- PRECAST(PRECASTObj, K = K)
  
  seuInt <- PRECASTObj@seulist[[1]]
  seuInt@meta.data$cluster <- factor(unlist(PRECASTObj@resList[[1]]$cluster))
  seuInt@meta.data$batch <- 1
  seuInt <- Add_embed(PRECASTObj@resList[[1]]$hZ[[1]], seuInt, embed_name = "PRECAST")
  posList <- lapply(PRECASTObj@seulist, function(x) cbind(x$row, x$col))
  seuInt <- Add_embed(posList[[1]], seuInt, embed_name = "position")
  Idents(seuInt) <- factor(seuInt@meta.data$cluster)
  #return(seuInt@reductions$PRECAST@cell.embeddings)
  return(seuInt$cluster)
}

drsc_wrapper_cluster <- function(exprs, coords, pct0 = 0.999, K = 7) {
  adj <- create_adjacency(coords = coords, pct0 = pct0)
  drsc_obj <- DR.SC_fit(X = exprs, Adj_sp = adj, K = K)
  lowdim <- drsc_obj$Objdrsc[[1]]$hZ
  return(drsc_obj$Objdrsc[[1]]$cluster)
}

wsir_cluster <- function(exprs, coords, K, slices = 8, alpha = 4) {
  wsir_out <- wSIR(X = exprs, 
                   coords = coords,
                   slices = slices,
                   alpha = alpha)
  wsir_kmns_out <- kmeans(x = wsir_out$scores, centers = K)
  return(wsir_kmns_out$cluster)
}

spca_wrapper_cluster <- function(counts = NULL, # 1) These need to be genes x cells (standard counts slot sce)
                         # 2) These need to have colnames (cell names)
                         coords = NULL,
                         ncores = 10,
                         K = 7,
                         gene.number=3000,
                         gene.type = "spatial") {
  spca_res <- CreateSpatialPCAObject(counts = counts, # create SPCA object
                                     location = coords,
                                     numCores_spark = ncores,
                                     gene.number = gene.number,
                                     gene.type = gene.type)
  spca_res <- SpatialPCA_buildKernel(spca_res, # build kernel (from spca tutorial)
                                     kerneltype="gaussian", 
                                     bandwidthtype="SJ",
                                     bandwidth.set.by.user=NULL,
                                     sparseKernel = TRUE,
                                     sparseKernel_ncore = ncores)
  spca_res <- SpatialPCA_EstimateLoading(spca_res, # estimate loadings (spca tutorial)
                                         fast=TRUE,
                                         SpatialPCnum=20) 
  spca_res <- SpatialPCA_SpatialPCs(spca_res, fast=TRUE) # calculate spatialPCs (spca tutorial)
  spca_projected <- t(spca_res@SpatialPCs) # extract projected matrix
  kept_cells <- spca_res@normalized_expr %>% colnames() # spca removes some cells - check which are left
  kept_indices <- colnames(counts) %in% kept_cells # find which of original cells are kept
  coords_kept <- coords[kept_indices,] # subset original coords to just get the coords of cells spca kept
  #return(list(projected = spca_projected, coords = coords_kept)) # return projected and (subsetted) coords
  return(list(cluster = walktrap_clustering(clusternum=K,
                                       latent_dat=spca_res@SpatialPCs,knearest=70),
              spots = colnames(spca_res@normalized_expr)))
}
```

```{r}
wsir_aris <- c()
precast_aris <- c()
drsc_aris <- c()
spca_aris <- c()

for (current_sample in samples_vec) {
  # extract this sample
  spe_DLPFC_currentsample <- spe_DLPFC[,spe_DLPFC@colData$sample == current_sample]
  
  # filter HVGs
  modelled_var_v1 <- modelGeneVar(spe_DLPFC_currentsample@assays@data$logcounts)
  topgenes_v1 <- scran::getTopHVGs(modelled_var_v1, n = 500)
  spe_DLPFC_currentsample <- spe_DLPFC_currentsample[topgenes_v1,
                                                     !is.na(spe_DLPFC_currentsample$layer_guess_reordered_short)]
  
  # extract current exprs, coords, counts
  exprs1 <- t(spe_DLPFC_currentsample@assays@data$logcounts)
  coords1 <- spe_DLPFC_currentsample@int_colData@listData$spatialCoords
  counts1 <- t(spe_DLPFC_currentsample@assays@data$counts)
  domains1 <- spe_DLPFC_currentsample@colData$layer_guess_reordered_short
  
  current_k <- length(unique(spe_DLPFC_currentsample@colData$layer_guess_reordered_short))

  # create clusters
  wsir_clusters <- wsir_cluster(exprs = exprs1,
                                coords = coords1,
                                K = current_k)
  precast_clusters <- precast_wrapper_cluster(exprs = exprs1,
                                              coords = coords1,
                                              K = current_k)
  drsc_clusters <- drsc_wrapper_cluster(exprs = exprs1,
                                        coords = coords1,
                                        K = current_k)
  spca_clusters_obj <- spca_wrapper_cluster(counts = t(counts1),
                                            coords = coords1,
                                            K = current_k)
  
  # extract spatial domains for precast and spca
  precast_cells <- spe_DLPFC_currentsample@colData$spot_ID %in% names(precast_clusters)
  spca_cells <- spe_DLPFC_currentsample@colData$spot_ID %in% spca_clusters_obj$spots
  precast_domains <- domains1[precast_cells]
  spca_domains <- domains1[spca_cells]
  domain_list <- list(domains1, 
                      precast_domains, 
                      domains1, 
                      spca_domains)
  wsir_aris <- wsir_aris %>% append(ARI(domain_list[[1]], wsir_clusters))
  precast_aris <- precast_aris %>% append(ARI(domain_list[[2]], precast_clusters))
  drsc_aris <- drsc_aris %>% append(ARI(domain_list[[3]], as.vector(drsc_clusters)))
  spca_aris <- spca_aris %>% append(ARI(domain_list[[4]], spca_clusters_obj$cluster))
}
```

STAMP

```{r}
library(aricode)
samples_vec <- unique(spe_DLPFC@colData$sample)
ari_dlpfc_stamp <- c()
for (current_sample in samples_vec) {
  # extract this sample
  spe_DLPFC_currentsample <- spe_DLPFC[,spe_DLPFC@colData$sample == current_sample]
  
  # filter HVGs
  modelled_var_v1 <- modelGeneVar(spe_DLPFC_currentsample@assays@data$logcounts)
  topgenes_v1 <- scran::getTopHVGs(modelled_var_v1, n = 500)
  spe_DLPFC_currentsample <- spe_DLPFC_currentsample[topgenes_v1,]
  
  # extract current coords and counts
  coords1 <- spe_DLPFC_currentsample@int_colData@listData$spatialCoords
  counts1 <- t(spe_DLPFC_currentsample@assays@data$counts)
  domains1 <- spe_DLPFC_currentsample@colData$layer_guess_reordered_short

  stamp_out <- stamp_wrapper_complete(X = counts1,
                                      coords = as.matrix(coords1))
  
  stamp_cells <- domains1[(spe_DLPFC_currentsample@colData$spot_ID %in% rownames(stamp_out$coords))]
  notna <- !is.na(stamp_cells)

  current_k <- length(unique(spe_DLPFC_currentsample@colData$layer_guess_reordered_short))
  
  stamp_kmns_out <- kmeans(x = stamp_out$embedding, centers = current_k)
  current_clusters <- stamp_kmns_out$cluster
  
  current_ari <- ARI(current_clusters[notna], stamp_cells[notna])
  
  ari_dlpfc_stamp <- ari_dlpfc_stamp %>% append(current_ari)
}
```

analyse

```{r}
aris_vec <- c(wsir_aris, precast_aris, drsc_aris, spca_aris)
aris_vec <- aris_vec %>% append(ari_dlpfc_stamp)
```

