---
title: "analysis_supp_fig1"
author: '500455892'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

analysis for supp fig 1 time benchmarking

```{r}
source("comparison_functions2.R")
source("spatial_benchmark_functions.R")
library(tidyverse)
library(PRECAST)
library(SeuratObject)
library(wSIR)
library(DR.SC)
library(ape)
library(vctrs)
library(SpatialPCA)
library(bluster)
library(doBy)
library(bench)
```

load data. See manuscript data availability section to download xenium breast cancer data sample 1. Will then need to adjust file paths.

```{r}
xb1_exprs <- readRDS(file = "/enna/nobackup/maximilw/xb1_exprs.Rds")
xb1_coords <- readRDS(file = "/enna/nobackup/maximilw/xb1_coords.Rds") %>% as.data.frame()
xb1_counts <- readRDS(file = "/enna/nobackup/maximilw/xb1_counts.Rds")
```

Create matrix where column i contains indices of points to take when want to use a certain number of points. E.g first we want to take 1000 points, so column 1 contains randomly 1000 1s and 0 everywhere else.

```{r}
sizes_to_take <- c(1,5,10,20,50,100,160)*1000
take_mat_time <- matrix(NA, nrow = nrow(xb1_coords), ncol = length(sizes_to_take))

i <- 1
for (current_size in sizes_to_take) {
  rows_to_take <- sample(x = c(1:nrow(xb1_coords)), size = current_size)
  take_mat_time[,i] <- c(1:nrow(xb1_coords)) %in% rows_to_take
  i <- i + 1
}
save(take_mat_time, file = "take_mat_time.RData")
```

STAMP

prep functions since STAMP is python. Will need to set up your own python to make this run.
```{r}
use_python("/Library/Frameworks/Python.framework/Versions/3.8/bin/python3")

test_file <- "stamp_wrapper.py"
source_python(test_file)

stamp_wrapper_complete <- function(X, coords) {
  out <- stamp_wrapper_fn(counts = as.matrix(X), 
                          coordinates = coords)
  stamp_cells_taken <- as.numeric(out[[2]]) + 1 # +1 since python 0-indexed
  stamp_coords_taken <- coords[stamp_cells_taken,]
  return(list(embedding = out[[1]],
              coords = stamp_coords_taken))
}
```

run stamp time analysis

```{r}
library(bench)
sizes_to_take <- c(1,5,10,20,50,100,160)*1000
load("take_mat_time.RData")
stamp_mems <- c()
stamp_times <- c()

for (i in 1:(ncol(take_mat_time)-3)) {
  print(paste0("Current size: ", sizes_to_take[i]))
  rows_to_take <- take_mat_time[,i]
  current_coords <- xb1_coords[rows_to_take,]
  current_counts <- xb1_counts[rows_to_take,]
  
  # STAMP
  stamp_mem <- mark(stamp_out <- stamp_wrapper_complete(X = current_counts, 
                                                    coords = as.matrix(current_coords)))
  stamp_mems[i] <- stamp_mem$mem_alloc
  stamp_times[i] <- stamp_mem$total_time
}
save(stamp_mems, stamp_times, file = "stamp_time_res.RData")
```


Other methods

```{r}
wsir_mems <- c()
wsir_times <- c()
pca_mems <- c()
pca_times <- c()
lda_mems <- c()
lda_times <- c()
drsc_mems <- c()
drsc_times <- c()
precast_mems <- c()
precast_times <- c()
for (i in 1:ncol(take_mat_time)) {
  rows_to_take <- take_mat_time[,i]
  current_coords <- xb1_coords[rows_to_take,]
  current_exprs <- xb1_exprs[rows_to_take,]
  
  # wSIR
  wsir_mem <- mark(wsir_out <- wSIR(X = current_exprs,
                                    coords = current_coords))
  wsir_mems[i] <- wsir_mem$mem_alloc
  wsir_times[i] <- wsir_mem$total_time
  
  # PCA
  pca_mem <- mark(pca_out <- prcomp(current_exprs))
  pca_mems[i] <- pca_mem$mem_alloc
  pca_times[i] <- pca_mem$total_time
  
  # LDA
  lda_mem <- mark(lda_out <- lda_compare(exprs_train = current_exprs,
                                               exprs_test = current_exprs,
                                               coords_train = current_coords,
                                               slices = 10))
  lda_mems[i] <- lda_mem$mem_alloc
  lda_times[i] <- lda_mem$total_time

  # DRSC
  drsc_mem <- mark(drsc_out <- drsc_wrapper(exprs = current_exprs,
                                                  coords = as.matrix(current_coords)))
  drsc_mems[i] <- drsc_mem$mem_alloc
  drsc_times[i] <- drsc_mem$total_time
  
  # PRECAST
  precast_mem <- mark(precast_out <- precast_wrapper(exprs = current_exprs,
                                                           coords = current_coords))
  precast_mems[i] <- precast_mem$mem_alloc
  precast_times[i] <- precast_mem$total_time
}

comp_time_df <- data.frame(time = c(wsir_times, pca_times, lda_times, drsc_times, precast_times),
                           memory = c(wsir_mems, pca_mems, lda_mems, drsc_mems, precast_mems),
                           method = vec_rep_each(c("wSIR", "PCA", "LDA", "DRSC", "PRECAST"), 7),
                           size = rep(sizes_to_take,5))
load("stamp_time_res.RData")
stamp_time_res_df <- data.frame(time = stamp_times,
                                memory = stamp_mems,
                                method = rep("STAMP", 4),
                                size = sizes_to_take[1:4])
comp_time_df <- rbind(comp_time_df, stamp_time_res_df)
```

PLS (since it wasn't included the first run)

```{r}
load("time_benchmark/take_mat_time.RData")
sizes_to_take <- c(1,5,10,20,50,100,150)*1000

pls_mems <- c()
pls_times <- c()
for (i in 1:ncol(take_mat_time)) {
  print(paste0("Current size: ", sizes_to_take[i]))
  rows_to_take <- take_mat_time[,i]
  current_coords <- xb1_coords[rows_to_take,]
  current_exprs <- xb1_exprs[rows_to_take,]
  
  # PLS
  pls_mem <- mark(pls_out <- pls_compare(exprs_train = current_exprs,
                                         exprs_test = current_exprs,
                                         coords_train = current_coords,
                                         coords_test = current_coords))
  pls_mems[i] <- pls_mem$mem_alloc
  pls_times[i] <- pls_mem$total_time
}

pls_resdf <- data.frame(time = pls_times,
                        memory = pls_mems,
                        method = rep("PLS", length(sizes_to_take)),
                        size = sizes_to_take)
comp_time_df <- rbind(comp_time_df, pls_resdf)
```

SpatialPCA

```{r}
load("time_benchmark/take_mat_time.RData")
sizes_to_take <- c(1,5,10,20,50,100,150)*1000

spca_mems <- c()
spca_times <- c()
for (i in 1:(ncol(take_mat_time)-4)) {
  print(paste0("Current size: ", sizes_to_take[i]))
  rows_to_take <- take_mat_time[,i]
  current_coords <- xb1_coords[rows_to_take,]
  current_exprs <- xb1_exprs[rows_to_take,]
  current_counts <- xb1_counts[rows_to_take,]
  rownames(current_counts) <- rownames(current_coords)
  
  # SpatialPCA
  spca_mem <- mark(spca_out <- spca_wrapper(counts = t(as.matrix(current_counts)),
                                            coords = as.matrix(current_coords),
                                            ncores = 1))
  spca_mems[i] <- spca_mem$mem_alloc
  spca_times[i] <- spca_mem$total_time
}
save(c(spca_mems, spca_times), file = "time_benchmark/spca_comp_res.RData")

spca_resdf <- data.frame(time = spca_times,
                         memory = spca_mems,
                         method = rep("SpatialPCA", 3),
                         size = sizes_to_take[1:3])
comp_time_df <- rbind(comp_time_df, spca_resdf)
```

```{r}
save(comp_time_df, file = "comp_time_df.RData")
```

