---
title: "create_figs"
output: html_document
date: "2025-12-18"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This document contains code to generate wSIR revision figures.

```{r}
library(tidyverse)
library(vctrs)
```

# Supplementary figure 1 - robustness to tile specification

1A

```{r}
load("data/response_fig_data/bcdcs_vec.RData")
rotation_vals <- c(0:8) * 10
rotate_bcdc_df <- data.frame(bcdc = bcdcs_vec,
                             angle = rep(rotation_vals, 10))
rotate_coords_fig <- ggplot(rotate_bcdc_df, aes(x = as.factor(angle), y = bcdc)) +
  geom_boxplot(colour = "black") +
  theme_classic() +
  labs(x = "Rotation Angle",
       y = "BCDC") +
  theme(text = element_text(size = 15))
rotate_coords_fig
ggsave(plot = rotate_coords_fig, filename = "figures/supp_fig1/rotate_coords_fig.pdf")
```

1B

```{r}
load("data/response_fig_data/shift_bcdc_vals.RData")
shift_amounts <- c(0:9)/10
bcdc_shift_df <- data.frame(bcdc = bcdc_vals,
                            shift_amount = rep(shift_amounts, 10))
shift_coords_fig <- ggplot(bcdc_shift_df, aes(x = as.factor(shift_amount), y = bcdc)) +
  geom_boxplot(colour = "black") +
  labs(x = expression(delta),
       y = "BCDC") +
  theme_classic() +
  theme(text = element_text(size = 15))
shift_coords_fig
ggsave(plot = shift_coords_fig, filename = "figures/supp_fig1/shift_coords_fig.pdf")
```

# Supplementary figure 2 - multiple sample improvement

```{r}
load("response_figs/response_fig_data/bcdcs_multisample.RData")
load("response_figs/response_fig_data/nsim_means_multisample.RData")
load("response_figs/response_fig_data/nsim_sds_multisample.RData")
test_sets_bcdc <- c() # each test_set loop, 3 bcdc values are createdd
test_sets_nsim <- c() # each loop, 3*length(n_prop) nsim values are created
train_sets_bcdc <- c()
train_sets_nsim <- c()
nsizes_nsim <- c()
for (test_set in c(1:3)) {
  # find indices for both test sets given the training set index
  train1 <- (test_set %% 3) + 1
  train2 <- ((test_set+1) %% 3) + 1
  current_train_vec <- c(paste0(train1, ",", train2), train1, train2)
  train_sets_bcdc <- train_sets_bcdc %>% append(current_train_vec)
  test_sets_bcdc <- test_sets_bcdc %>% append(rep(test_set, 3))
}
```

2A

```{r}
bcdcs_means <- c()
bcdcs_sds <- c()
for (i in 1:9) { # want to have each element of bcdcs_means be mean of i, i+9, i+18, ... 
  current_vals <- bcdcs_multisample[seq(from = i, by = 9, length.out = 10)]
  bcdcs_means[i] <- mean(current_vals)
  bcdcs_sds[i] <- sd(current_vals)
}
bcdcs_bs_df <- data.frame(mean = bcdcs_means,
                          sd = bcdcs_sds,
                          train = train_sets_bcdc, # run line ~150 for this # this is now in prev chunk
                          test = test_sets_bcdc)
bcdcs_bs_df$ntrain <- as.factor(rep(c(2,1,1), 3))
custom_order = c("1,2", "3,1", "2,3", "1", "2", "3")
multisample_bcdc_fig <- ggplot(bcdcs_bs_df, aes(x = factor(train, levels = custom_order), 
                        y = mean,
                        fill = ntrain)) +
  facet_wrap(~test, scales = "free_x") +
  geom_col(width = 0.6, colour = "black") +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.4) +
  theme_classic() +
  labs(x = "Training set", 
       y = "Bias corrected distance correlation") +
  theme(legend.position = "none",
        text = element_text(size = 15)) +
  scale_fill_manual(values = c("2" = "blue3", "1" = "lightblue"))
multisample_bcdc_fig
ggsave(plot = multisample_bcdc_fig, filename = "response_figs/supp_fig2/multisample_bcdc_fig.pdf")
```

2B

```{r}
nsim_bs_means <- c()
nsim_bs_sds <- c()
n_props <- c(0.005, 0.01, 0.05, 0.1, 0.2)

for (i in 1:45) {
  current_means <- nsim_means_multisample[seq(from = i, by = 45, length.out = 10)]
  nsim_bs_means[i] <- mean(current_means)
  nsim_bs_sds[i] <- sd(current_means)
}
nsim_bs_df <- data.frame(mean = nsim_bs_means,
                         sd = nsim_bs_sds,
                         test_set = vec_rep_each(c(1,2,3),15),
                         props = rep(n_props, 9),
                         train_set = as.factor(vec_rep_each(train_sets_bcdc,5)))
multisample_nsim_fig <- ggplot(nsim_bs_df, aes(x = props, y = mean, colour = train_set)) +
  geom_line() +
  facet_wrap(~test_set, scales = "free_y") +
  theme_classic() +
  labs(x = "Proportion of test set that is neighbourhood size",
       y = "Mean jaccard value",
       colour = "Training set") +
  scale_colour_manual(values = c("1" = "lightblue", 
                                 "2" = "lightblue",
                                 "3" = "lightblue",
                                 "1,2" = "blue3",
                                 "2,3" = "blue3",
                                 "3,1" = "blue3")) +
  geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd, fill = train_set), 
              alpha = 0.2,
              colour = NA) +
  scale_fill_manual(values = c("1" = "lightblue", 
                                 "2" = "lightblue",
                                 "3" = "lightblue",
                                 "1,2" = "blue3",
                                 "2,3" = "blue3",
                                 "3,1" = "blue3")) +
  guides(fill = "none") +
  theme(text = element_text(size = 15),
        legend.position = "bottom")
multisample_nsim_fig
ggsave(plot = multisample_nsim_fig, filename = "response_figs/supp_fig2/multisample_nsim_fig.pdf")
```

# Supplementary figure 3 - thymus trajectory case study

```{r}
load("response_figs/response_fig_data/sce.sling_pca.RData")
load("response_figs/response_fig_data/sce.sling.RData")
load("response_figs/response_fig_data/sce_merged.RData")
load("response_figs/response_fig_data/umap_pca_hm.RData")
load("response_figs/response_fig_data/umap_wsir_hm.RData")
load("response_figs/response_fig_data/thymus_sample_sizes.RData")
```

B, C

```{r}
traj_umap_wsir_pst_plot <- ggplot(data.frame(umapwsir1_hm = umap_wsir_hm$layout[,1],
                  umapwsir2_hm = umap_wsir_hm$layout[,2],
                  cma = sce_merged$cma_v2,
                  pseudotime = sce.sling$slingPseudotime_1),
       aes(x = umapwsir1_hm, y = umapwsir2_hm, colour = cma)) + 
  geom_point() +
  theme_classic() + 
  labs(x = "UMAP1 on harmonised wSIR", 
       y = "UMAP2 on harmonised wSIR",
       colour = "CMA") +
  ggtitle("UMAP on wSIR post-harmony coloured by CMA") +
  theme(legend.position = "bottom")
ggsave(plot = traj_umap_wsir_pst_plot, filename = "response_figs/supp_fig3/traj_umap_wsir_pst_plot.pdf")
traj_umap_wsir_cma_plot <- ggplot(data.frame(umapwsir1_hm = umap_wsir_hm$layout[,1],
                  umapwsir2_hm = umap_wsir_hm$layout[,2],
                  cma = sce_merged$cma_v2,
                  pseudotime = sce.sling$slingPseudotime_1),
       aes(x = umapwsir1_hm, y = umapwsir2_hm, colour = pseudotime)) + 
  geom_point() +
  theme_classic() + 
  labs(x = "UMAP1 on harmonised wSIR", 
       y = "UMAP2 on harmonised wSIR",
       colour = "Pseudotime") +
  ggtitle("UMAP on wSIR post-harmony coloured by slingshot pseudotime") +
  theme(legend.position = "bottom")
ggsave(plot = traj_umap_wsir_cma_plot, filename = "response_figs/supp_fig3/traj_umap_wsir_cma_plot.pdf")
```

D, E

```{r}
traj_umap_pca_pst_plot <- ggplot(data.frame(umappca1_hm = umap_pca_hm$layout[,1],
                  umappca2_hm = umap_pca_hm$layout[,2],
                  cma = sce_merged$cma_v2,
                  pseudotime = sce.sling$slingPseudotime_1),
       aes(x = umappca1_hm, y = umappca2_hm, colour = cma)) + 
  geom_point() +
  theme_classic() + 
  labs(x = "UMAP1 on harmonised PCA", 
       y = "UMAP2 on harmonised PCA",
       colour = "CMA") +
  ggtitle("UMAP on PCA post-harmony coloured by CMA") +
  theme(legend.position = "bottom")
ggsave(plot = traj_umap_pca_pst_plot, filename = "response_figs/supp_fig3/traj_umap_pca_pst_plot.pdf")
traj_umap_pca_cma_plot <- ggplot(data.frame(umappca1_hm = umap_pca_hm$layout[,1],
                  umappca2_hm = umap_pca_hm$layout[,2],
                  cma = sce_merged$cma_v2,
                  pseudotime = sce.sling$slingPseudotime_1),
       aes(x = umappca1_hm, y = umappca2_hm, colour = pseudotime)) + 
  geom_point() +
  theme_classic() + 
  labs(x = "UMAP1 on harmonised PCA", 
       y = "UMAP2 on harmonised PCA",
       colour = "Pseudotime") +
  ggtitle("UMAP on PCA post-harmony coloured by slingshot pseudotime") +
  theme(legend.position = "bottom")
ggsave(plot = traj_umap_pca_cma_plot, filename = "response_figs/supp_fig3/traj_umap_pca_cma_plot.pdf")
```

A preparation

```{r}
sce.sling1 <- sce.sling[,1:n1]
sce.sling2 <- sce.sling[,(n1+1):(n1+n2)]
sce.sling3 <- sce.sling[,(n1+n2+1):(n1+n2+n3)]
sce.sling4 <- sce.sling[,(n1+n2+n3+1):(n1+n2+n3+n4)]
sce.sling5 <- sce.sling[,(n1+n2+n3+n4+1):(n1+n2+n3+n4+n5)]
sce.sling6 <- sce.sling[,(n1+n2+n3+n4+n5+1):(n1+n2+n3+n4+n5+n6)]
sce.sling7 <- sce.sling[,(n1+n2+n3+n4+n5+n6+1):(n1+n2+n3+n4+n5+n6+n7)]
sce.sling8 <- sce.sling[,(n1+n2+n3+n4+n5+n6+n7+1):(n1+n2+n3+n4+n5+n6+n7+n8)]
sce.sling9 <- sce.sling[,(n1+n2+n3+n4+n5+n6+n7+n8+1):(n1+n2+n3+n4+n5+n6+n7+n8+n9)]

sce.sling_pca1 <- sce.sling_pca[,1:n1]
sce.sling_pca2 <- sce.sling_pca[,(n1+1):(n1+n2)]
sce.sling_pca3 <- sce.sling_pca[,(n1+n2+1):(n1+n2+n3)]
sce.sling_pca4 <- sce.sling_pca[,(n1+n2+n3+1):(n1+n2+n3+n4)]
sce.sling_pca5 <- sce.sling_pca[,(n1+n2+n3+n4+1):(n1+n2+n3+n4+n5)]
sce.sling_pca6 <- sce.sling_pca[,(n1+n2+n3+n4+n5+1):(n1+n2+n3+n4+n5+n6)]
sce.sling_pca7 <- sce.sling_pca[,(n1+n2+n3+n4+n5+n6+1):(n1+n2+n3+n4+n5+n6+n7)]
sce.sling_pca8 <- sce.sling_pca[,(n1+n2+n3+n4+n5+n6+n7+1):(n1+n2+n3+n4+n5+n6+n7+n8)]
sce.sling_pca9 <- sce.sling_pca[,(n1+n2+n3+n4+n5+n6+n7+n8+1):(n1+n2+n3+n4+n5+n6+n7+n8+n9)]
```

A

```{r}
traj1_plot <- plotReducedDim(sce.sling1, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca1, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling1, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj2_plot <- plotReducedDim(sce.sling2, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca2, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling2, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj3_plot <- plotReducedDim(sce.sling3, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca3, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling3, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj4_plot <- plotReducedDim(sce.sling4, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca4, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling4, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj5_plot <- plotReducedDim(sce.sling5, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca5, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling5, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj6_plot <- plotReducedDim(sce.sling6, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca6, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling6, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj7_plot <- plotReducedDim(sce.sling7, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca7, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling7, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj8_plot <- plotReducedDim(sce.sling8, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca8, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling8, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")

traj9_plot <- plotReducedDim(sce.sling9, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "wSIR Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling_pca9, "spatial", colour_by = "slingPseudotime_1") + 
  labs(title = "PCA Pseudotime", x = "Spatial 1", y = "Spatial 2") + 
  theme(legend.title = element_blank(), legend.position="top") +
  plotReducedDim(sce.sling9, "spatial", colour_by = "cma_v2") + 
  labs(title = "CMA axis", x = "Spatial 1", y = "Spatial 2") +
  theme(legend.title = element_blank(), legend.position="top")
```

```{r}
ggsave(plot = traj1_plot, filename = "response_figs/supp_fig3/traj1_plot.pdf")
ggsave(plot = traj2_plot, filename = "response_figs/supp_fig3/traj2_plot.pdf")
ggsave(plot = traj3_plot, filename = "response_figs/supp_fig3/traj3_plot.pdf")
ggsave(plot = traj4_plot, filename = "response_figs/supp_fig3/traj4_plot.pdf")
ggsave(plot = traj5_plot, filename = "response_figs/supp_fig3/traj5_plot.pdf")
ggsave(plot = traj6_plot, filename = "response_figs/supp_fig3/traj6_plot.pdf")
ggsave(plot = traj7_plot, filename = "response_figs/supp_fig3/traj7_plot.pdf")
ggsave(plot = traj8_plot, filename = "response_figs/supp_fig3/traj8_plot.pdf")
ggsave(plot = traj9_plot, filename = "response_figs/supp_fig3/traj9_plot.pdf")
```

# Supplementary figure 4 - DLPFC data for spatial domains

```{r}
load("response_figs/response_fig_data/dplfc_psR2_curve_df.RData")

dlpfc_colours <- c("SpatialPCA" = "orangered3",
                   "PRECAST" = "darkgreen",
                   "DR.SC" = "green3",
                   "STAMP" = "orange2",
                   "wSIR" = "blue")
dlpfc_psR2_curve_plot <- ggplot(dplfc_psR2_curve_df, aes(x = index, y = mean, colour = method, group = method)) +
  geom_line() +
  geom_ribbon(aes(ymin = mean - sd, ymax = mean + sd, colour = NULL, fill = method), alpha = 0.2) +
  theme_classic() +
  scale_colour_manual(values = dlpfc_colours) +
  scale_fill_manual(values = dlpfc_colours) +
  labs(x = "Number of components of low-dimensional embedding", y = "Pseudo-R2", colour = "DR method", fill = "DR method") +
  theme(legend.position = "bottom")
ggsave(plot = dlpfc_psR2_curve_plot, filename = "response_figs/supp_fig4/dlpfc_psR2_curve_plot.pdf")
```

```{r}
load("response_figs/response_fig_data/aris_vec.RData")
load("response_figs/response_fig_data/samples_vec.RData")
aris_res_df <- data.frame(ari = aris_vec,
                          method = vec_rep_each(c("wSIR", "PRECAST", "DR.SC", "SpatialPCA", "STAMP"),
                                                length(samples_vec)),
                          sample = rep(samples_vec,5))
ari_boxplot <- ggplot(aris_res_df, aes(x = method, y = ari, colour = method)) +
  geom_boxplot() +
  scale_colour_manual(values = dlpfc_colours) +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 15)) +
  labs(x = "Method", y = "ARI")
ari_boxplot
ggsave(plot = ari_boxplot, filename = "response_figs/supp_fig4/ari_boxplot.pdf")
```

# Supplementary figure 5 - computational requirements benchmarking

```{r}
load("response_figs/response_fig_data/comp_time_df.RData")
computation_colours <- c("PRECAST" = "darkgreen",
                         "DRSC" = "green3",
                         "STAMP" = "orange2",
                         "wSIR" = "blue",
                         "LDA" = "lightgrey",
                         "PCA" = "burlywood1",
                         "PLS" = "darkgrey",
                         "SpatialPCA" = "orangered3")
options("scipen"=10) # this makes x axis show 100000 instead of 1e5
computation_time_fig <- comp_time_df %>% ggplot(aes(x = as.factor(size), y = log(time), colour = method, group = method)) +
  geom_point() +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  labs(x = "Dataset size",
       y = "Computational time ln(seconds)",
       colour = "") +
  geom_line() +
  scale_colour_manual(values = computation_colours)
computation_time_fig
ggsave(plot = computation_time_fig, filename = "response_figs/supp_fig5/computation_time_fig.pdf")

computation_memory_fig <- comp_time_df %>% ggplot(aes(x = as.factor(size), y = log(memory), colour = method, group = method)) +
  geom_point() +
  theme_classic() +
  theme(text = element_text(size = 15),
        legend.position = "bottom") +
  labs(x = "Dataset size",
       y = "Memory usage ln(bytes)",
       colour = "") +
  geom_line() +
  scale_colour_manual(values = computation_colours)
computation_memory_fig
ggsave(plot = computation_memory_fig, filename = "response_figs/supp_fig5/computation_memory_fig.pdf")
```


# Supplementary figure 6 - further results on Tangram experiment

```{r}
load("response_figs/response_fig_data/jsd_allsizes_em1.RData")
fig4_colours <- c("Counts" = "green3",
                  "LogCounts" = "green4",
                  "LDA" = "lightgrey",
                  "PCA" = "grey",
                  "PLS" = "darkgrey",
                  "SIR" = "lightblue",
                  "wSIR" = "blue")
n_props <- c(0.005,0.01,0.05,0.1,0.2)
jsd_em1_res_df <- data.frame(mean = c(jsd_allsizes_em1[[1]]$mean,
                                      jsd_allsizes_em1[[2]]$mean,
                                      jsd_allsizes_em1[[3]]$mean,
                                      jsd_allsizes_em1[[4]]$mean,
                                      jsd_allsizes_em1[[5]]$mean),
                             sd = c(jsd_allsizes_em1[[1]]$sd,
                                    jsd_allsizes_em1[[2]]$sd,
                                    jsd_allsizes_em1[[3]]$sd,
                                    jsd_allsizes_em1[[4]]$sd,
                                    jsd_allsizes_em1[[5]]$sd),
                             method = rep(c("LDA", "wSIR", "SIR", "PCA", "PLS", "Counts", "LogCounts"),5),
                             n_size = vec_rep_each(n_props, 7))
min_jsd_df <- jsd_em1_res_df %>% 
  group_by(n_size) %>% 
  summarize(min_value = min(mean)) %>% 
  as.data.frame()
min_jsd_df$method <- jsd_em1_res_df %>% group_by(n_size) %>%
  slice_min(mean) %>%
  pull(method)

jsd_em1_plot <- jsd_em1_res_df %>% ggplot(aes(x = method, y = mean, colour = method)) + 
  geom_point() +
  facet_wrap(~n_size, scales = "free", nrow = 1) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd, x = method)) +
  theme_classic() +
  scale_colour_manual(values = fig4_colours) +
  theme(text = element_text(size = 12),
        axis.text.x = element_blank(),
        legend.position = "bottom") +
  labs(y = "Jensen-Shannon Divergence",
       x = "",
       colour = "Tangram input") +
  geom_hline(data = min_jsd_df, aes(yintercept = min_value, color = method), linetype = "dashed") +
  ggtitle("Jensen-Shannon Divergence between true and predicted coordinates")
ggsave(plot = jsd_em1_plot, filename = "response_figs/supp_fig6/jsd_em1_plot.pdf")
```

```{r}
load("response_figs/response_fig_data/all_rank_cors_df.RData")
spearmancor_em1_plot <- ggplot(all_rank_cors_df, aes(x = method, y = rank_cors, colour = method)) +
  geom_boxplot() +
  scale_colour_manual(values = fig4_colours) +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 15)) +
  labs(x = "Method", y = "Spearman rank correlation")
ggsave(plot = spearmancor_em1_plot, filename = "response_figs/supp_fig6/spearmancor_em1_plot.pdf")
```


# Supplementary figure 7 - analysis of PRECAST simulated dataset

Nsim

```{r}
dlpfc_colours <- c("SpatialPCA" = "orangered3",
                   "PRECAST" = "darkgreen",
                   "DR.SC" = "green3",
                   "STAMP" = "orange2",
                   "wSIR" = "blue")
load("response_figs/response_fig_data/precast_sim_nsim_df.RData")
precastsim_nsim_plot <- ggplot(precast_sim_nsim_df, aes(x = n_size, y = nsim, colour = method, group = method)) +
  geom_line() +
  facet_wrap(~sample) +
  scale_colour_manual(values = dlpfc_colours) +
  theme_classic() +
  theme(legend.position = "bottom") +
  labs(x = "Neighbourhood size as percentage of test set",
       y = "Jaccard index",
       colour = "DR method") +
  theme(text = element_text(size = 15))
ggsave(plot = precastsim_nsim_plot, filename = "drsc_simulation/precastsim_nsim_plot.pdf")
```

BCDC

```{r}
benchmark_colours <- c("SpatialPCA" = "lightgrey",
                       "PRECAST" = "grey",
                       "DR.SC" = "darkgrey",
                       "STAMP" = "grey1",
                       "wSIR" = "blue")
load("response_figs/response_fig_data/precast_sim_bcdc_df.RData")
precastsim_bcdc_plot <- ggplot(precast_sim_bcdc_df, aes(x = method, y = bcdc, fill = method)) +
  geom_col(colour = "black", width = 0.7) +
  facet_wrap(~sample) +
  theme_classic() +
  theme(legend.position = NULL,
        axis.text.x = element_text(angle = 45, hjust = 1)) +
  scale_fill_manual(values = benchmark_colours) +
  theme(legend.position = "none",
        text = element_text(size = 15)) +
  labs(x = "Method", y = "Bias-corrected distance correation")
ggsave(plot = precastsim_bcdc_plot, filename = "drsc_simulation/precastsim_bcdc_plot.pdf")
```

Pseudo-R2 curves

```{r}
load("response_figs/response_fig_data/precastsim_psr2_df.RData")
dlpfc_colours <- c("SpatialPCA" = "orangered3",
                   "PRECAST" = "darkgreen",
                   "DR.SC" = "green3",
                   "STAMP" = "orange2",
                   "wSIR" = "blue")
precatsim_psR2_plot1 <- ggplot(precastsim_psr2_df, aes(x = ncomp, y = psR2_s1, colour = method)) +
  geom_line() +
  theme_classic() + 
  labs(x = "Number of components", y = "Pseudo-R2", colour = "DR method") +
  scale_colour_manual(values = dlpfc_colours) +
  theme(legend.position = "bottom",
        text = element_text(size = 15))
precatsim_psR2_plot2 <- ggplot(precastsim_psr2_df, aes(x = ncomp, y = psR2_s2, colour = method)) +
  geom_line() +
  theme_classic() + 
  labs(x = "Number of components", y = "Pseudo-R2", colour = "DR method") +
  scale_colour_manual(values = dlpfc_colours) +
  theme(legend.position = "bottom",
        text = element_text(size = 15))
precatsim_psR2_plot3 <- ggplot(precastsim_psr2_df, aes(x = ncomp, y = psR2_s3, colour = method)) +
  geom_line() +
  theme_classic() + 
  labs(x = "Number of components", y = "Pseudo-R2", colour = "DR method") +
  scale_colour_manual(values = dlpfc_colours) +
  theme(legend.position = "bottom",
        text = element_text(size = 15))
ggsave(plot = precatsim_psR2_plot1, filename = "drsc_simulation/precastsim_psR2_plot1.pdf")
ggsave(plot = precatsim_psR2_plot2, filename = "drsc_simulation/precastsim_psR2_plot2.pdf")
ggsave(plot = precatsim_psR2_plot3, filename = "drsc_simulation/precastsim_psR2_plot3.pdf")
```


UMAPs, spatial plots, figure legend

```{r}
library(Seurat)
load("response_figs/response_fig_data/umaps_out_s1.RData")
load("response_figs/response_fig_data/umaps_out_s2.RData")
load("response_figs/response_fig_data/umaps_out_s3.RData")
githubURL <- "https://github.com/feiyoung/PRECAST/blob/main/vignettes_data/data_simu.rda?raw=true"
download.file(githubURL,"data_simu.rda",mode='wb')
load("data_simu.rda")
# sample 1
precast_sim1_coords <- cbind(data_simu[[1]]$row, data_simu[[1]]$col)
data_simu[[1]]
precast_sim1_counts <- data_simu[[1]]@assays$RNA$counts
precast_sim1_exprs <- t(LogNormalize(precast_sim1_counts))
# sample 2
precast_sim2_coords <- cbind(data_simu[[2]]$row, data_simu[[2]]$col)
precast_sim2_counts <- data_simu[[2]]@assays$RNA$counts
precast_sim2_exprs <- t(LogNormalize(precast_sim2_counts))
# sample 3
precast_sim3_coords <- cbind(data_simu[[3]]$row, data_simu[[3]]$col)
precast_sim3_counts <- data_simu[[3]]@assays$RNA$counts
precast_sim3_exprs <- t(LogNormalize(precast_sim3_counts))

# save sample sizes
n1 <- nrow(precast_sim1_coords)
n2 <- nrow(precast_sim2_coords)
n3 <- nrow(precast_sim3_coords)
# save spatial domains
precast_sim1_domains <- data_simu[[1]]$true_cluster
precast_sim2_domains <- data_simu[[2]]$true_cluster
precast_sim3_domains <- data_simu[[3]]$true_cluster
```

```{r}
precastsim_full_umap_df <- data.frame(x = c(unlist(lapply(umaps_out_s1, function(x)return(x[,1]))),
                                            unlist(lapply(umaps_out_s2, function(x)return(x[,1]))),
                                            unlist(lapply(umaps_out_s3, function(x)return(x[,1])))),
                                      y = c(unlist(lapply(umaps_out_s1, function(x)return(x[,2]))),
                                            unlist(lapply(umaps_out_s2, function(x)return(x[,2]))),
                                            unlist(lapply(umaps_out_s3, function(x)return(x[,2])))),
                                      domain = c(rep(precast_sim1_domains,5),
                                                 rep(precast_sim2_domains,5),
                                                 rep(precast_sim3_domains,5)),
                                      sample = c(rep("Sample 1", n1*5),
                                                 rep("Sample 2", n2*5),
                                                 rep("Sample 3", n3*5)),
                                      method = c(vec_rep_each(c("wSIR", "DR.SC", "PRECAST", "SpatialPCA", "STAMP"), n1),
                                                 vec_rep_each(c("wSIR", "DR.SC", "PRECAST", "SpatialPCA", "STAMP"), n2),
                                                 vec_rep_each(c("wSIR", "DR.SC", "PRECAST", "SpatialPCA", "STAMP"), n3)))
precastsim_full_umap_df$method_sample <- paste0(precastsim_full_umap_df$method, " ", precastsim_full_umap_df$sample)

umap_plot_full <- ggplot(precastsim_full_umap_df, aes(x = x, y = y, colour = as.factor(domain))) +
  geom_point(size = 0.5) +
  theme_classic() +
  facet_wrap(~method_sample, nrow = 5, scales = "free") +
  labs(x = "UMAP 1",
       y = "UMAP 2") +
  theme(legend.position = "none",
        text = element_text(size = 12))
ggsave(plot = umap_plot_full, filename = "drsc_simulation/umap_plot_full.pdf", width = 6, height = 10)
```

```{r}
precastsim_spatialplot <- ggplot(data.frame(x = c(precast_sim1_coords[,1],
                                                  precast_sim2_coords[,1],
                                                  precast_sim3_coords[,1]),
                                            y = c(precast_sim1_coords[,2],
                                                  precast_sim2_coords[,2],
                                                  precast_sim3_coords[,2]),
                                            domain = c(precast_sim1_domains,
                                                       precast_sim2_domains,
                                                       precast_sim3_domains),
                                            sample = c(rep("Sample 1", n1),
                                                       rep("Sample 2", n2),
                                                       rep("Sample 3", n3))),
                                  aes(x = x, y = y, colour = as.factor(domain))) +
  geom_point(size = 0.5) +
  theme_classic() +
  theme(legend.position = "none",
        text = element_text(size = 12)) +
  labs(x = "Spatial 1", y = "Spatial 2") +
  facet_wrap(~sample)
ggsave(plot = precastsim_spatialplot, filename = "drsc_simulation/precastsim_spatialplot.pdf", width = 6, height = 2)
```

```{r}
library(cowplot)
legend_domains <- get_legend(ggplot(data.frame(x = umaps_out_s1[[1]][,1],
                  y = umaps_out_s1[[1]][,2],
                  domain = precast_sim1_domains),
       aes(x = x, y = y, colour = as.factor(domain))) +
  geom_point() +
  theme_classic() +
  labs(x = "UMAP 1",
       y = "UMAP 2",
       colour = "Spatial domain") +
  theme(legend.position = "bottom"))
domain_legend_plot <- plot_grid(legend_domains)
ggsave(plot = domain_legend_plot, filename = "drsc_simulation/domain_legend_plot.pdf")
```

# Benchmarking vs other spatial DR methods

Xenium Breast Cancer

```{r}
benchmark_colours <- c("SpatialPCA" = "orangered3",
                       "PRECAST" = "darkgreen",
                       "DR.SC" = "green3",
                       "STAMP" = "orange2",
                       "wSIR" = "blue")
load("response_figs/response_fig_data/nsim_xb1_df.RData")
load("response_figs/response_fig_data/bcdc_xb1_df.RData")
load("response_figs/response_fig_data/nsim_xb2_df.RData")
load("response_figs/response_fig_data/bcdc_xb2_df.RData")

nsim_xball_df <- rbind(nsim_xb1_df, nsim_xb2_df)
nsim_xball_df$sample <- vec_rep_each(c("Sample 1", "Sample 2"),20)

benchmark_colours <- c("PRECAST" = "lightgrey",
                       "DR.SC" = "darkgrey",
                       "STAMP" = "grey2",
                       "wSIR" = "blue")
nsim_xball_plot <- ggplot(nsim_xball_df, aes(x = 100*n_prop, y = means, colour = method)) + 
  geom_line() +
  facet_wrap(~sample) +
  theme_classic() +
  theme(legend.position = "bottom",
        text = element_text(size = 15)) +
  geom_ribbon(aes(x = 100*n_prop, ymin = means - sds, ymax = means + sds, fill = method),
              colour = NA, alpha = 0.2) +
  labs(x = "Percent of dataset used as neighbourhood size",
       colour = "DR method",
       fill = "DR method",
       y = "Mean jaccard index value") +
  scale_fill_manual(values = benchmark_colours) + 
  scale_colour_manual(values = benchmark_colours)
nsim_xball_plot
ggsave(nsim_xball_plot, filename = "benchmark_xb1/nsim_xball_plot.pdf")
```

```{r}
bcdc_xball_df <- rbind(bcdc_xb1_df, bcdc_xb2_df)
bcdc_xball_df$sample <- vec_rep_each(c("Sample 1", "Sample 2"),4)

bcdc_xball_plot <- ggplot(bcdc_xball_df, aes(x = reorder(method, mean), 
                                         y = mean, 
                                         fill = method)) +
  geom_col(width = 0.8) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.5) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5)) +
  facet_wrap(~sample) +
  labs(x = "DR method", y = "Bias-corrected distance correlation") +
  scale_fill_manual(values = benchmark_colours) +
  theme(legend.position = "",
        text = element_text(size = 15))
bcdc_xball_plot
ggsave(bcdc_xball_plot, filename = "benchmark_xb1/bcdc_xball_plot.pdf")
```

Embryo

```{r}
load("response_figs/response_fig_data/bcdc_all_df.RData")
bcdc_all_plot <- ggplot(bcdc_all_df, aes(x = reorder(method, mean), 
                                         y = mean, 
                                         fill = method)) +
  geom_col(width = 0.8) +
  geom_errorbar(aes(ymin = mean - sd, ymax = mean + sd), width = 0.5) +
  theme_classic() + 
  theme(axis.text.x = element_text(angle = 45, vjust = 0.5, hjust=0.5)) +
  facet_wrap(~sample) +
  labs(x = "DR method", y = "Bias-corrected distance correlation") +
  scale_fill_manual(values = benchmark_colours) +
  theme(legend.position = "")
bcdc_all_plot
```

```{r}
load("response_figs/response_fig_data/nsim_df.RData")

benchmark_colours <- c("SpatialPCA" = "grey",
                       "PRECAST" = "lightgrey",
                       "DR.SC" = "darkgrey",
                       "STAMP" = "grey2",
                       "wSIR" = "blue")
nsim_all_plot <- ggplot(nsim_df, aes(x = 100*n_prop, y = means, colour = method)) + 
  geom_line() +
  facet_wrap(~sample) +
  theme_classic() +
  theme(legend.position = "bottom") +
  geom_ribbon(aes(x = 100*n_prop, ymin = means - sds, ymax = means + sds, fill = method),
              colour = NA, alpha = 0.2) +
  labs(x = "Percent of dataest used as neighbourhood size",
       colour = "DR method",
       fill = "DR method",
       y = "Mean jaccard index value") +
  scale_fill_manual(values = benchmark_colours) + 
  scale_colour_manual(values = benchmark_colours)
nsim_all_plot
ggsave(nsim_all_plot, filename = "benchmark_embeddings/nsim_all_plot.pdf")
```

# Reviewer figure 1 - null spatial experiment via shuffling rows

```{r}
load("response_figs/response_fig_data/bcdc_shuffle_df.RData")
shuffle_bcdc_plot <- ggplot(bcdc_shuffle_df, aes(x = shuffle, y = bcdc)) + 
  geom_boxplot() +
  theme_classic() +
  theme(text = element_text(size = 15)) +
  labs(x = "Shuffle rows",
       y = "Bias-corrected distance correlation")
shuffle_bcdc_plot
ggsave(plot = shuffle_bcdc_plot, filename = "response_figs/reviewer_fig1/shuffle_bcdc_plot.pdf")
```

```{r}
load("response_figs/response_fig_data/evalues_summary_df.RData")
shuffle_evalues_plot <- ggplot(evalues_summary_df[evalues_summary_df$index<11,], aes(x = index, y = mean, colour = shuffle)) +
  geom_line() +
  theme_classic() +
  geom_ribbon(aes(ymin = mean - sd, 
                  ymax = mean + sd, 
                  fill = shuffle, 
                  x = index), 
              colour = NA,
              alpha = 0.5) +
  geom_hline(yintercept = 0) + 
  theme(text = element_text(size = 15)) +
  labs(x = "Low-dimensional component", y = "Eigenvalue", colour = "Shuffle", fill = "Shuffle") +
  theme(legend.position = "bottom") +
  scale_x_continuous(breaks = c(1:10))
shuffle_evalues_plot
ggsave(plot = shuffle_evalues_plot, filename = "response_figs/reviewer_fig1/shuffle_evalues_plot.pdf")
```


