---
title: "plots_main_fig4"
author: '500455892'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Plots for main fig 4: exploration of ductal breast cancer dataset.

Note: there is no analysis .rmd file for this figure since all the analysis is inside this document.

```{r}
library(SingleCellExperiment)
library(scater)
library(StabMap)
library(zellkonverter)
library(wSIR)
library(patchwork)
library(BiocParallel)
library(bluster)
library(pheatmap)
library(scran)
library(dplyr)
library(ggrepel)
library(ggplotify)

set.seed(2024)
```

Note the dataset is too large to include in this github repo. To access the dataset, please refer to the data availability section of the manuscript.

```{r}
subset_lower.tri = function(m) {
  mm = m[lower.tri(m, diag = FALSE)]
  return(c(mm))
}

spe_raw <- readH5AD("../../data/xenium_breast_cancer/xenium_rep1.h5ad")
spe_raw <- addPerCellQCMetrics(spe_raw, assay.type = "X")
spe_raw <- spe_raw[, spe_raw$total >= 10 & spe_raw$detected >= 5]

spe_raw <- logNormCounts(spe_raw, exprs_values = "X")

ct = "DCIS_1"

spe <- spe_raw[,spe_raw$Cluster == ct]
# spe <- spe_raw

spe <- addPerFeatureQCMetrics(spe, assay.type = "X")

spe <- runPCA(spe)
spe <- runUMAP(spe)

exprs_raw = t(assay(spe, "logcounts"))
coords_raw = as.data.frame(reducedDim(spe, "spatial"))

out = wSIR(X = exprs_raw,
           coords = coords_raw,
           slices = 30,
           alpha = 4,
           maxDirections = 30,
           varThreshold = 0.99,
           optim_params = FALSE)

varimportance = getVarImportance(out$directions)

reducedDim(spe, "wSIR") <- out[[1]]

spe <- runUMAP(spe, dimred = "wSIR", name = "UMAP_wSIR", BPPARAM = MulticoreParam(workers = 4))

pts = 0

set.seed(2024)
m = clusterRows(reducedDim(spe, "wSIR"),
                BLUSPARAM = TwoStepParam(first = KmeansParam(centers = 1000, iter.max = 10, nstart = 5),
                                         second = NNGraphParam(cluster.fun = "louvain", k = 20)))
spe$m <- m

set.seed(2024)
mp = clusterRows(reducedDim(spe, "PCA"),
                 BLUSPARAM = TwoStepParam(first = KmeansParam(centers = 1000, iter.max = 10, nstart = 5),
                                          second = NNGraphParam(cluster.fun = "louvain", k = 20)))
spe$mp <- mp
pcaimportance = getVarImportance(attr(reducedDim(spe, "PCA"), "rotation"))[rownames(spe)]
```

# Plot A

For this section, the dataset is too large to include in this github repo. To access the data, please see the data availability section of the manuscript.

Preparation
```{r}
xen <- readH5AD("../../data/xenium_breast_cancer/xenium_rep1.h5ad")
xen <- addPerCellQCMetrics(xen, assay.type = "X")
xen <- xen[, xen$total >= 10 & xen$detected >= 5]
xen <- logNormCounts(xen, exprs_values = "X")

library(DropletUtils)

spat = read.csv("../../data/DCIS_Visium/spatial/tissue_positions.csv", header = TRUE, row.names = 1)

# read in the visium annotation
annot = read.delim("../../data/DCIS_Visium/Cell_Barcode_Type_Matrices_Visium.tsv", row.names = 1)


spe = read10xCounts("../../data/DCIS_Visium/filtered_feature_bc_matrix")
colnames(spe) <- spe$Barcode
rownames(spe) <- make.unique(rowData(spe)$Symbol)
reducedDim(spe, "spatial") <- as.data.frame(as.matrix(spat[spe$Barcode, c("array_row", "array_col")]) %*% matrix(c(1,0, 0, 1),2, 2))

spe$Annotation <- annot[colnames(spe), "Annotation"]

spe <- logNormCounts(spe)

spe <- addPerFeatureQCMetrics(spe)
spe <- spe[rowData(spe)$detected > 5,]

genes = intersect(rownames(spe), rownames(xen))

spe <- spe[genes,]

out = calculatewSIR(x = spe,
                    dimred = "spatial",
                    slices = 20,
                    alpha = 4,
                    maxDirections = 30,
                    varThreshold = 0.99,
                    optim_params = FALSE)

reducedDim(spe, "wSIR") <- out$scores

spe <- runUMAP(spe, dimred = "wSIR", name = "wSIR_UMAP")

spe <- runPCA(spe)
spe <- runUMAP(spe, dimred = "PCA", name = "PCA_UMAP")

spe <- addPerCellQCMetrics(spe)


spe$visium_cluster = clusterRows(reducedDim(spe, "wSIR"), BLUSPARAM = NNGraphParam(cluster.fun = "louvain", k = 20))
```

Plot
```{r}
g = plotReducedDim(spe, "spatial", colour_by = "visium_cluster",point_shape = 16, point_alpha = 1, point_size = 2) + 
  theme(axis.text = element_blank()) +
  theme(axis.line = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 20)) +
  xlab("") + 
  ylab("") + 
  theme(aspect.ratio = 1) +
  # guides(colour = guide_legend(title = "", override.aes = list(size = 7)))
  theme(legend.position = "none") + 
  ggtitle("Visium DCIS")
```

# Plot B

```{r}
g = plotReducedDim(spe, "wSIR_UMAP", colour_by = "visium_cluster") + 
  theme(axis.text = element_blank()) +
  theme(axis.line = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 20)) +
  xlab("") + 
  ylab("") + 
  theme(aspect.ratio = 1) +
  guides(colour = guide_legend(title = "", override.aes = list(size = 6))) + 
  ggtitle("Visium DCIS")
```

# Plot C

```{r}
g = plotReducedDim(spe, "spatial", colour_by = "Annotation",point_shape = 16, point_alpha = 1, point_size = 2) + 
  theme(axis.text = element_blank()) +
  theme(axis.line = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 20)) +
  xlab("") + 
  ylab("") + 
  theme(aspect.ratio = 1) +
  guides(colour = guide_legend(title = "", override.aes = list(size = 7))) +
  theme(legend.position = "none") + 
  ggtitle("Visium DCIS")
```

# Plot D

```{r}
g = plotReducedDim(spe, "wSIR_UMAP", colour_by = "Annotation") + 
  theme(axis.text = element_blank()) +
  theme(axis.line = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 20)) +
  xlab("") + 
  ylab("") + 
  theme(aspect.ratio = 1) +
  # guides(colour = guide_legend(title = "", override.aes = list(size = 6))) + 
  ggtitle("Visium DCIS") + 
  theme(legend.position = "none")
```

# Plot E

```{r}
p = ComplexHeatmap::Heatmap(log10(unclass(table(spe$visium_cluster, spe$Annotation))+1), 
                            column_title = "Annotation",
                            row_title = "wSIR Clusters",
                            col = colorRampPalette(c("white", "orange", "red", "black"))(100),
                            heatmap_legend_param = list(title = "Number of\nspots\n(log10)")
)

g = as.ggplot(p)
```

# Plot H

Note the plots are no longer in alphabetical order. This is so the code can be run in the order it appears in this document.

```{r}
library(Rfast)

proj = wSIR::projectWSIR(wsir = out, newdata = t(assay(xen, "logcounts")[genes,]))

reducedDim(xen, "wSIR_proj") <- proj

xen <- runPCA(xen)
xen <- runUMAP(xen, dimred = "PCA", name = "PCA_UMAP")
xen <- runUMAP(xen, dimred = "wSIR_proj", name = "wSIR_proj_UMAP")

g = plotReducedDim(xen, "wSIR_proj_UMAP", colour_by = "Cluster") +
  theme(axis.text = element_blank()) +
  theme(axis.line = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 20)) +
  xlab("") + 
  ylab("") + 
  theme(aspect.ratio = 1) +
  guides(colour = guide_legend(title = "", override.aes = list(size = 7))) + 
  ggtitle("Xenium DCIS mapped onto Visium embedding") +
  NULL
```

# Plot J

```{r}
xen$tmp <- ifelse(xen$Cluster == "Macrophages_1", "mac", "Not")
xen$selected <- ifelse(reducedDim(xen, "wSIR_proj_UMAP")[,1] > 0 &
                         reducedDim(xen, "wSIR_proj_UMAP")[,1] < 4 &
                         reducedDim(xen, "wSIR_proj_UMAP")[,2] > 3, "selected", "not")

xen$mac_sub <- ifelse(xen$Cluster == "Macrophages_1", "Macrophages_1", "Other")
xen$mac_sub[xen$selected == "selected"] <- "Macrophages_1 subset"

g = plotReducedDim(xen[,rev(order(xen$mac_sub))], "spatial", colour_by = "mac_sub",
                   point_size = 0.5) + 
  theme(axis.text = element_blank()) +
  theme(axis.line = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 20)) +
  xlab("") + 
  ylab("") + 
  theme(aspect.ratio = 1) +
  # guides(colour = guide_legend(title = "", override.aes = list(size = 7)))
  theme(legend.position = "bottom") + 
  # scale_colour_manual(values = setNames(c("black", "grey90"), c("selected", "not")))
  scale_colour_manual(values = setNames(c("black", "grey90", "blue"), c("Macrophages_1 subset", "Other", "Macrophages_1"))) + 
  guides(colour = guide_legend(title = "", override.aes = list(size = 15), theme = theme(legend.text = element_text(size = 15))))
```

# Plot I

```{r}
mk = findMarkers(xen,
                 groups = droplevels(factor(xen$selected)),
                 direction = "any")
DE_df = cbind(mk[["selected"]], gene = rownames(mk[["selected"]]))
g = ggplot(DE_df, aes(x = summary.logFC, y = -log10(p.value+1e-100))) + 
  geom_point(colour = "grey") + 
  geom_text_repel(aes(label = gene), data = DE_df[DE_df$Top <= 10 & DE_df$logFC.not > 0,]) + 
  xlab("Log Fold Change") + 
  ylab("-log(P-Value)") + 
  # ggtitle(paste0("wSIR Cluster ", clusts[1],
  #                " vs Cluster ", clusts[2])) + 
  theme_classic() + 
  ggtitle("Macrophages_1 subset marker genes")
```


# Plot F

```{r}
varimportance = getVarImportance(out$directions)
pcaimportance = getVarImportance(attr(reducedDim(spe, "PCA"), "rotation"))

importance_df = data.frame(PCA = pcaimportance[genes], wSIR = varimportance[genes], gene = genes)

g = ggplot(importance_df, aes(x = PCA, y = wSIR)) + 
  theme_classic() +
  geom_point(colour = "grey") + 
  geom_text_repel(aes(label = gene), data = subset(importance_df, PCA > 0.7 | wSIR > 0.3 | PCA+wSIR > 0.6),
                  size = 2) +
  xlab("PCA gene relevance") +
  ylab("wSIR gene relevance") +
  theme(axis.title = element_text(size = 10)) +
  # theme(axis.line = element_blank()) +
  theme(axis.ticks = element_blank()) +
  theme(plot.title = element_text(hjust = 0.5, size = 20)) +
  # xlab("") + 
  # ylab("") + 
  theme(aspect.ratio = 1) +
  NULL
```

# Plot G

```{r}
g = wrap_plots(sapply(c("MMP12", "OPRPN", "POSTN", "PTGDS", "SFRP4"), function(gene) {
  plotReducedDim(spe, "spatial", colour_by = gene, point_shape = 16, point_alpha = 1, point_size = 1.35) + 
    guides(colour = guide_legend(override.aes = list(size = 10), title = "")) + 
    ggtitle(gene) +
    theme(axis.text = element_blank()) +
    theme(axis.line = element_blank()) +
    theme(axis.ticks = element_blank()) +
    theme(plot.title = element_text(hjust = 0.5)) +
    xlab("") + 
    ylab("") + 
    theme(aspect.ratio = 1) +
    theme(legend.position = "none")
}, simplify = FALSE), nrow = 1)
```








