---
title: "analysis_main_fig2"
author: '500455892'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(PRECAST)
library(SeuratObject)
library(wSIR)
library(DR.SC)
library(vctrs) # for vec_rep_each function
library(SpatialPCA)
library(bluster)
library(doBy)
source("spatial_benchmark_functions.R")
source("wrappers.R")
```


Analysis for main fig 2: donut simulation, parameter robustness, benchmarking wSIR vs spatial methods

# Donut simulation

Note this contains the code to perform the analysis and create the plot. 

```{r}
# pls function
library(FNN)
library(Matrix)
library(pls)
library(tidyverse)

pls_wrap = function(X, coords, ...) {
  # return the scores from pls where
  # X is a cells x genes expression matrix (explanatory variables)
  # Y is a cells x response(s) matrix
  # ... addition arguments to plsr, e.g. ncomp may be wanted to identify the 
  # number of components to return
  
  plsr_model <- plsr(coords ~ X, ...)
  
  return(list(plsr_model$scores,
              plsr_model$loadings,
              plsr_model$ncomp,
              plsr_model
  ))
  
}

n = 1000
set.seed(200)
rvals = runif(n, 0.1, 1)
thetavals = runif(n, -pi, pi) + 0.1


xvals = rvals*cos(thetavals) 
yvals = rvals*sin(thetavals)

# create some genes that are uncorrelated with space but have high variation
p_nonspatial = 10
nonspatialgenes = matrix(rnorm(n*p_nonspatial, 0, 5), p_nonspatial, n)
# nonspatialgenes = t(matrix(rnorm(n*p_nonspatial, nonspvals, 5), n, p_nonspatial))
#dim(nonspatialgenes)

# create some genes that are correlated with space but have low variation
p_r = 5
spatialgenes_r = t(matrix(rnorm(n*p_r, rvals, 0.01), n, p_r))
#dim(spatialgenes_r)

p_th = 5
spatialgenes_t = t(matrix(rnorm(n*p_th, thetavals, 0.5), n, p_th))
#dim(spatialgenes_t)

# put together the data
X = t(rbind(nonspatialgenes, spatialgenes_r, spatialgenes_t))

cn <- c()
for (i in 1:10) {
  cn <- cn %>% append(paste0("non_spatial_gene_", i))
}
for (i in 1:5) {
  cn <- cn %>% append(paste0("radius_gene_", i))
}
for (i in 1:5) {
  cn <- cn %>% append(paste0("angle_gene_", i))
}
colnames(X) <- cn
coords = data.frame(x = xvals, y = yvals)

### Graph PCA
# Number of neighbors
k <- 6
# Get k-nearest neighbors
knn_result <- get.knn(coords, k = k)
# Create the adjacency matrix (sparse format)
adj_matrix <- sparseMatrix(
  i = rep(1:nrow(coords), each = k),
  j = as.vector(knn_result$nn.index),
  x = 1
)
# Create the degree matrix (diagonal matrix)
degree_matrix <- Diagonal(x = rowSums(adj_matrix))
# Compute the Laplacian matrix
laplacian_matrix <- degree_matrix - adj_matrix
# Convert to full matrix if needed
graphL <- as.matrix(laplacian_matrix)
lambda <- 0.5
G <- Diagonal(n) + lambda*graphL
Ginv <- Rfast::spdinv(as.matrix(G))
Expr <- scale(X)
C <- t(Expr) %*% Ginv %*% Expr
eigenC <- eigen(C, symmetric = T)
out_GPCA <- list()
out_GPCA$loadings <- eigenC$vectors[, 1:2]
out_GPCA$scores <- Ginv %*% Expr %*% out_GPCA$loadings

out = wSIR:::wSIRSpecifiedParams(X, 
                                 coords, 
                                 varThreshold = 0.95,
                                 alpha = 4,
                                 slices = 6)
out_pc = princomp(X)
out_pls = pls_wrap(X, as.matrix(coords))

library(MASS)
tile_lda = wSIR:::spatialAllocator(coords, slices = 6)[,3]
mod_lda = lda(tile_lda~., data = as.data.frame(X))
out_lda <- predict(mod_lda, as.data.frame(X))
out_lda$scores <- out_lda$x

df_all <- data.frame(coords,
                     wSIR = out$scores[, 1:2],
                     PLS = out_pls[[1]][, 1:2],
                     PC = out_pc$scores[, 1:2], 
                     GPCA = out_GPCA$scores[, 1:2],
                     LDA = out_lda$scores[,1:2],
                     thetavals = thetavals, rvals = rvals) %>%
  mutate(across(3:12, scale)) %>%
  mutate(wSIR.2 = -wSIR.2, wSIR.1 = -wSIR.1) %>%
  pivot_longer(cols = wSIR.1:rvals, names_to = "dir", values_to = "value") %>%
  mutate(dir = factor(dir, levels = c("thetavals", "wSIR.1", "GPCA.1", "LDA.LD1", "PLS.Comp.1","PC.Comp.1",
                                      "rvals", "wSIR.2", "GPCA.2", "LDA.LD2", "PLS.Comp.2", "PC.Comp.2"),
                      labels = c("Angle", "wSIR1", "GPCA1", "LDA1", "PLS1", "PC1",
                                      "Radius", "wSIR2", "GPCA2", "LDA2", "PLS2", "PC2")))
g = df_all %>% ggplot(aes(x = x, y = y, colour = abs(value))) +
  geom_point(size = 1.5) +
  facet_wrap(vars(dir), nrow = 2) +
  # scale_colour_viridis_c() +
  scale_colour_gradient(low = "#0077FF", high = "#FFCC00") +
  labs(x = "Y1", y = "Y2", colour = "") +
  theme_classic(base_size = 18 ) + 
  theme(axis.text.x =element_blank(),
        axis.ticks.x = element_blank(),
        axis.ticks.y = element_blank(),
        axis.text.y = element_blank()) + 
  theme(legend.position = "bottom") +
  theme(strip.background = element_rect(colour = NA)) +
  guides(colour = guide_colorbar(title = "Absolute score values")) +
  theme(aspect.ratio = 1)

ggsave(g, file = "donut_res_plot_color_abs_value.pdf", width = 14, height =  6)
```


# Parameter robustness

```{r}
load("MouseData.Rdata")
coords1 <- coords_list[[1]]
coords3 <- coords_list[[3]]
exprs1 <- exprs_logcount_list[[1]]
exprs3 <- exprs_logcount_list[[3]]
alpha_vals <- c(0,4,8,15,30,50)
slice_vals <- c(5,10,15,20,30,50)

# Define parameters
alpha_vals <- c(0, 4, 8, 15, 30, 50)
slice_vals <- c(5, 10, 15, 20, 30, 50)
nrep <- 5

# Precompute lengths
n_alpha <- length(alpha_vals)
n_slices <- length(slice_vals)

# Preallocate the bc_dcs vector
total_iterations <- n_alpha * n_slices * nrep
bc_dcs <- numeric(total_iterations)

# Parallelized loop using mclapply
results <- lapply(seq_along(alpha_vals), function(a_idx) {
  alpha <- alpha_vals[a_idx]
  cat(sprintf("Current alpha: %d\n", alpha))
  
  sapply(seq_along(slice_vals), function(s_idx) {
    slices <- slice_vals[s_idx]
    cat(sprintf("  Current slices: %d\n", slices))
    
    replicate(nrep, {
      # Random sampling
      rsample1 <- sample(c(TRUE, FALSE), replace = TRUE, size = nrow(coords1), prob = c(0.8, 0.2))
      rsample3 <- sample(c(TRUE, FALSE), replace = TRUE, size = nrow(coords3), prob = c(0.8, 0.2))
      
      # Subsetting
      exprs_train <- exprs3[rsample3, ]
      coords_train <- coords3[rsample3, ]
      exprs_test <- exprs1[rsample1, ]
      coords_test <- coords1[rsample1, ]
      
      # Create wSIR object and calculate bc_dcs
      wsir_obj <- wSIR:::wSIRSpecifiedParams(X = exprs_train,
                                             coords = coords_train,
                                             alpha = alpha,
                                             slices = slices,
                                             varThreshold = 0.95)
      projected <- projectWSIR(wsir = wsir_obj, newdata = exprs_test)
      bc_distCor(projected, coords_test)
    })
  })
})

# Flatten the results into the bc_dcs vector

```

```{r}
# each column is one alpha, and each row is one slice
bc_dcs_avg <- sapply(results, colMeans)
bc_dc_params_train3test1_df <- data.frame(bc_distcor = c(bc_dcs_avg),
                                          alpha = factor(rep(alpha_vals, each = 6)),
                                          slices = factor(rep(slice_vals, 6)))

ggplot(bc_dc_params_train3test1_df, aes(x = alpha, y = slices, fill = bc_distcor)) +
  geom_tile() +
  scale_fill_viridis_c()

saveRDS(bc_dc_params_train3test1_df, file = "embryo_results_bc_dc_params_train3test1_df.Rds")
```

Xenium breast cancer

First need to preprocess the data

```{r setup, include=FALSE}
library(reticulate)
use_python("/Library/Frameworks/Python.framework/Versions/3.12/bin/python3")
source_python("graphPCA_wrap.py")
knitr::opts_chunk$set(echo = TRUE)
source("comparison_functions2.R")
```

Note the xenium data files are too large for github. To run this code, please see the data availability section of the paper and download the data from there. 

```{r, eval=F}
# Step 1: Load the .h5ad file
spe <- anndata::read_h5ad("../../data/xenium_rep1.h5ad")

# Step 2: Add per-cell QC metrics
# Assuming raw counts are stored in "X"
spe$obs$total <- rowSums(spe$X)
spe$obs$detected <- rowSums(spe$X > 0)

# Step 3: Filter cells based on QC metrics
filtered_cells <- spe$obs$total >= 10 & spe$obs$detected >= 5
spe <- spe[filtered_cells, ]

# Step 4: Examine cluster composition and subset to "Stromal" cells
cluster_counts <- table(spe$obs$Cluster) # Cluster composition
sorted_cluster_counts <- sort(cluster_counts) # Sorted cluster composition
stromal_cells <- spe$obs$Cluster == "Stromal" # Logical vector for "Stromal" cells
spe <- spe[stromal_cells, ]

counts_raw <- spe$X # Save the raw counts data for graphPCA
spe2 <- AnnData2SCE(spe)
spe2 <- logNormCounts(spe2, exprs_values = "X")

# Step 6: Save raw expression data and spatial coordinates
exprs_raw = t(assay(spe2, "logcounts"))
coords_raw = as.data.frame(reducedDim(spe2, "spatial"))
save(counts_raw, exprs_raw, coords_raw, file = "xenium_data.Rdata")
```

Begin analysis

```{r}
load("xenium_data.Rdata")

# Define parameters
alpha_vals <- c(0, 4, 8, 15, 30, 50)
slice_vals <- c(5, 10, 15, 20, 30, 50)
nrep <- 5

# Precompute lengths
n_alpha <- length(alpha_vals)
n_slices <- length(slice_vals)

# Preallocate the bc_dcs vector
total_iterations <- n_alpha * n_slices * nrep
bc_dcs <- numeric(total_iterations)

# Parallelized loop using mclapply
results <- lapply(seq_along(alpha_vals), function(a_idx) {
  alpha <- alpha_vals[a_idx]
  cat(sprintf("Current alpha: %d\n", alpha))
  
  sapply(seq_along(slice_vals), function(s_idx) {
    slices <- slice_vals[s_idx]
    cat(sprintf("  Current slices: %d\n", slices))
    
    replicate(nrep, {
      # Random sampling
      rsample1 <- sample(c(TRUE, FALSE), replace = TRUE, size = nrow(coords_raw), prob = c(0.8, 0.2))
      
      # Subsetting
      exprs_train <- exprs_raw[rsample1, ]
      coords_train <- coords_raw[rsample1, ]
      exprs_test <- exprs_raw[!rsample1, ]
      coords_test <- coords_raw[!rsample1, ]
      
      # Create wSIR object and calculate bc_dcs
      wsir_obj <- wSIR:::wSIRSpecifiedParams(X = exprs_train,
                                             coords = coords_train,
                                             alpha = alpha,
                                             slices = slices,
                                             varThreshold = 0.95)
      projected <- wSIR:::projectWSIR(wsir = wsir_obj, newdata = exprs_test)
      bc_distCor(projected, coords_test)
    })
  })
})

# Flatten the results into the bc_dcs vector

```

```{r}
# each column is one alpha, and each row is one slice
bc_dcs_avg <- sapply(results, colMeans)
bc_dc_params_train3test1_df <- data.frame(bc_distcor = c(bc_dcs_avg),
                                          alpha = factor(rep(alpha_vals, each = 6)),
                                          slices = factor(rep(slice_vals, 6)))

ggplot(bc_dc_params_df, aes(x = alpha, y = slices, fill = bc_distcor)) +
  geom_tile() +
  scale_fill_viridis_c()

saveRDS(bc_dc_params_df, file = "xenium_results_bc_dc_params.Rds")
```

# Benchmarking

## Embryo

download data

```{r}
# Packages needed to download data
library(scran) # for logNormCounts
library(MouseGastrulationData) # to download the data for this vignette


seqfish_data_sample1 <- LohoffSeqFISHData(samples = c(1,2))
seqfish_data_sample1 = logNormCounts(seqfish_data_sample1) # log transform variance stabilising
rownames(seqfish_data_sample1) <- rowData(seqfish_data_sample1)[,"SYMBOL"] # change rownames to gene symbols that are consistent
sample1_exprs = t(assay(seqfish_data_sample1, "logcounts")) # extract matrix of gene expressions
sample1_exprs_raw = t(assay(seqfish_data_sample1, "counts")) # extract counts of gene expressions
sample1_coords = spatialCoords(seqfish_data_sample1)[,1:2] %>% as.data.frame()
colnames(sample1_coords) = c("x", "y")

seqfish_data_sample2 <- LohoffSeqFISHData(samples = c(3,4))
seqfish_data_sample2 = logNormCounts(seqfish_data_sample2) # log transform variance stabilising
rownames(seqfish_data_sample2) <- rowData(seqfish_data_sample2)[,"SYMBOL"] # change rownames to gene symbols that are consistent
sample2_exprs = t(assay(seqfish_data_sample2, "logcounts")) # extract matrix of gene expressions
sample2_exprs_raw = t(assay(seqfish_data_sample2, "counts")) # extract counts of gene expressions
sample2_coords = spatialCoords(seqfish_data_sample2)[,1:2] %>% as.data.frame()
colnames(sample2_coords) = c("x", "y")

seqfish_data_sample3 <- LohoffSeqFISHData(samples = c(5,6))
seqfish_data_sample3 = logNormCounts(seqfish_data_sample3) # log transform variance stabilising
rownames(seqfish_data_sample3) <- rowData(seqfish_data_sample3)[,"SYMBOL"] # change rownames to gene symbols that are consistent
sample3_exprs = t(assay(seqfish_data_sample3, "logcounts")) # extract matrix of gene expressions
sample3_exprs_raw = t(assay(seqfish_data_sample3, "counts")) # extract counts of gene expressions
sample3_coords = spatialCoords(seqfish_data_sample3)[,1:2] %>% as.data.frame()
colnames(sample3_coords) = c("x", "y")
```

### BCDC

#### Sample 1

Create matrix of which cells to take at each iteration

```{r}
nrep <- 10
p_to_take <- 0.1

take_mat <- matrix(NA, nrow = nrow(sample1_coords), ncol = nrep) # column i is which rows to take in iteration i

for (i in 1:nrep) { # fill matrix of which rows to take each repeat
  to_take <- sample(x = c(TRUE, FALSE), size = nrow(sample1_coords), replace = TRUE, prob = c(p_to_take,(1-p_to_take)))
  take_mat[,i] <- to_take
}
#save(take_mat, file = "take_mat.RData")
```

Compute BCDCs

```{r}
load("take_mat.RData")
nrep <- dim(take_mat)[2]

bcdcs_wsir <- rep(0,nrep)
bcdcs_precast <- rep(0,nrep)
bcdcs_drsc <- rep(0,nrep)
bcdcs_spca <- rep(0,nrep)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat[,i]
  #current_counts <- sample1_counts[to_take,]
  current_coords <- sample1_coords[to_take,]
  current_ldo <- paste0("benchmark_embeddings/lowdim_out_r",i,".RData")
  load(current_ldo)
  current_spca_out <- paste0("benchmark_embeddings/spca_out_r",i,".RData")
  load(current_spca_out)
  bcdc_out <- metric_bcdc(low_dims = lowdim_out, coords = current_coords)
  bcdcs_wsir[i] <- bcdc_out$wSIR
  bcdcs_precast[i] <- bcdc_out$PRECAST
  bcdcs_drsc[i] <- bcdc_out$DRSC
  bcdcs_spca[i] <- energy::bcdcor(spca_out[[1]], spca_out[[2]])
}
```

STAMP

```{r}
load("take_mat.RData")

nrep <- dim(take_mat)[2]

bcdcs_stamp <- rep(0,nrep)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat[,i]
  current_counts <- sample1_counts[to_take,]
  current_coords <- sample1_coords[to_take,]
  stamp_out <- stamp_wrapper_complete(X = current_counts,
                                      coords = current_coords)
  bcdcs_stamp[i] <- energy::bcdcor(stamp_out[[1]], stamp_out[[2]])
}

#save(bcdcs_stamp, file = "bcdcs_stamp.RData")
```

Compile results

```{r}
#load("benchmark_embeddings/bcdcs_stamp.RData")
means <- c(mean(bcdcs_wsir),
                               mean(bcdcs_precast),
                               mean(bcdcs_drsc),
                               mean(bcdcs_spca),
                               mean(bcdcs_stamp))
sds <- c(sd(bcdcs_wsir),
                               sd(bcdcs_precast),
                               sd(bcdcs_drsc),
                               sd(bcdcs_spca),
                               sd(bcdcs_stamp))

bcdc_df <- data.frame(method = c("wSIR",
                                 "PRECAST",
                                 "DR.SC",
                                 "SpatialPCA",
                                 "STAMP"),
                      mean = means,
                      sd = sds
                      )
```

#### Sample 2

```{r}
nrep <- 10
p_to_take <- 0.1

take_mat2 <- matrix(NA, nrow = nrow(sample2_coords), ncol = nrep) # column i is which rows to take in iteration i

for (i in 1:nrep) { # fill matrix of which rows to take each repeat
  to_take <- sample(x = c(TRUE, FALSE), size = nrow(sample2_coords), replace = TRUE, prob = c(p_to_take,(1-p_to_take)))
  take_mat2[,i] <- to_take
}
#save(take_mat2, file = "take_mat2.RData")
```

```{r}
load("take_mat2.RData")
sample2_counts <- t(assay(seqfish_data_sample2, "counts"))

nrep <- dim(take_mat2)[2]

bcdcs_stamp2 <- rep(0,nrep)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat2[,i]
  current_counts <- sample2_counts[to_take,]
  current_coords <- as.matrix(sample2_coords)[to_take,]
  stamp_out <- stamp_wrapper_complete(X = current_counts,
                                      coords = current_coords)
  bcdcs_stamp2[i] <- energy::bcdcor(stamp_out[[1]], stamp_out[[2]])
}

#save(bcdcs_stamp2, file = "bcdcs_stamp2.RData")
```


```{r}
load("take_mat2.RData")
nrep <- dim(take_mat2)[2]

bcdcs_wsir <- rep(0,nrep)
bcdcs_precast <- rep(0,nrep)
bcdcs_drsc <- rep(0,nrep)
bcdcs_spca <- rep(0,nrep)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat2[,i]
  #current_counts <- sample2_counts[to_take,]
  current_coords <- sample2_coords[to_take,]
  current_ldo <- paste0("benchmark_embeddings2/lowdim_out_r",i,".RData")
  load(current_ldo)
  current_spca_out <- paste0("benchmark_embeddings2/spca_out_r",i,".RData")
  load(current_spca_out)
  bcdc_out <- metric_bcdc(low_dims = lowdim_out, coords = current_coords)
  bcdcs_wsir[i] <- bcdc_out$wSIR
  bcdcs_precast[i] <- bcdc_out$PRECAST
  bcdcs_drsc[i] <- bcdc_out$DRSC
  bcdcs_spca[i] <- energy::bcdcor(spca_out[[1]], spca_out[[2]])
}
```

```{r}
#load("benchmark_embeddings2/bcdcs_stamp2.RData")
means <- c(mean(bcdcs_wsir),
                               mean(bcdcs_precast),
                               mean(bcdcs_drsc),
                               mean(bcdcs_spca),
                               mean(bcdcs_stamp2))
sds <- c(sd(bcdcs_wsir),
                               sd(bcdcs_precast),
                               sd(bcdcs_drsc),
                               sd(bcdcs_spca),
                               sd(bcdcs_stamp2))

bcdc2_df <- data.frame(method = c("wSIR",
                                 "PRECAST",
                                 "DR.SC",
                                 "SpatialPCA",
                                 "STAMP"),
                      mean = means,
                      sd = sds
                      )
```


#### Sample 3

```{r}
nrep <- 10
p_to_take <- 0.1

take_mat3 <- matrix(NA, nrow = nrow(sample3_coords), ncol = nrep) # column i is which rows to take in iteration i

for (i in 1:nrep) { # fill matrix of which rows to take each repeat
  to_take <- sample(x = c(TRUE, FALSE), size = nrow(sample3_coords), replace = TRUE, prob = c(p_to_take,(1-p_to_take)))
  take_mat3[,i] <- to_take
}
#save(take_mat3, file = "take_mat3.RData")
```

```{r}
load("take_mat3.RData")
nrep <- dim(take_mat3)[2]

bcdcs_wsir <- rep(0,nrep)
bcdcs_precast <- rep(0,nrep)
bcdcs_drsc <- rep(0,nrep)
bcdcs_spca <- rep(0,nrep)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat3[,i]
  #current_counts <- sample3_counts[to_take,]
  current_coords <- sample3_coords[to_take,]
  current_ldo <- paste0("benchmark_embeddings3/lowdim_out_r",i,".RData")
  load(current_ldo)
  current_spca_out <- paste0("benchmark_embeddings3/spca_out_r",i,".RData")
  load(current_spca_out)
  bcdc_out <- metric_bcdc(low_dims = lowdim_out, coords = current_coords)
  bcdcs_wsir[i] <- bcdc_out$wSIR
  bcdcs_precast[i] <- bcdc_out$PRECAST
  bcdcs_drsc[i] <- bcdc_out$DRSC
  bcdcs_spca[i] <- energy::bcdcor(spca_out[[1]], spca_out[[2]])
}
```

```{r}
load("take_mat3.RData")
sample3_counts <- t(assay(seqfish_data_sample3, "counts"))

nrep <- dim(take_mat3)[2]

bcdcs_stamp3 <- rep(0,nrep)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat3[,i]
  current_counts <- sample3_counts[to_take,]
  current_coords <- as.matrix(sample3_coords)[to_take,]
  stamp_out <- stamp_wrapper_complete(X = current_counts,
                                      coords = current_coords)
  bcdcs_stamp3[i] <- energy::bcdcor(stamp_out[[1]], stamp_out[[2]])
}

#save(bcdcs_stamp3, file = "bcdcs_stamp3.RData")
```

```{r}
#load("benchmark_embeddings3/bcdcs_stamp3.RData")
means <- c(mean(bcdcs_wsir),
                               mean(bcdcs_precast),
                               mean(bcdcs_drsc),
                               mean(bcdcs_spca),
                               mean(bcdcs_stamp3))
sds <- c(sd(bcdcs_wsir),
                               sd(bcdcs_precast),
                               sd(bcdcs_drsc),
                               sd(bcdcs_spca),
                               sd(bcdcs_stamp3))

bcdc3_df <- data.frame(method = c("wSIR",
                                 "PRECAST",
                                 "DR.SC",
                                 "SpatialPCA",
                                 "STAMP"),
                      mean = means,
                      sd = sds
                      )
```

#### compile results of all samples

```{r}
bcdc_all_df <- rbind(bcdc_df, bcdc2_df, bcdc3_df)
bcdc_all_df$sample <- vec_rep_each(c("Embryo 1", "Embryo 2", "Embryo 3"), 5)
benchmark_colours <- c("SpatialPCA" = "grey",
                       "PRECAST" = "lightgrey",
                       "DR.SC" = "darkgrey",
                       "STAMP" = "grey2",
                       "wSIR" = "blue")
save(bcdc_all_df, file = "bcdc_all_df.RData")
```

### neighbourhood similarity

STAMP

```{r}
source("neighbourhood_similarity_functions.R")
jaccard_lowdim <- function(low_dims, # list of embeddings, e.g low_dims$wSIR contains wsir embedding
                           coords_list, # list of coords where each entry is the coords for the cells used in each dr method. e.g coords_list$spca is the coordinates of cells retained by spca
                           n_vals) { # n_vals is a list of neighbourhood sizes (e.g 5,10,20)
  true_dmats <- lapply(coords_list, dist) # create list of distance matrices
  embedding_dmats <- lapply(low_dims, dist)
  nmethods <- length(low_dims) # integer number of methods being compared
  results <- lapply(as.list(c(1:nmethods)), FUN = function(i) { # sapply over 1:nmethods so each time can extract low_dim[[i]], coords_lits[[i]] and dmats[[i]] more easily than loop over one of them
    embedding <- low_dims[[i]]
    coords <- coords_list[[i]]
    true_dmat <- as.matrix(true_dmats[[i]])
    embedding_dmat <- as.matrix(embedding_dmats[[i]])
    jaccard_n <- sapply(n_vals, FUN = function(n_size) { # for each neighbourhood size, compute (avg) jaccard value
      true_neighbours <- find_neighbours(dist_true = true_dmat, n = n_size) # compute n true neighbours
      current_jaccards <- jaccard_from_dmat(neighbours_true = true_neighbours, dist_dr = embedding_dmat)
      return(c(mean(current_jaccards), sd(current_jaccards)))
    })
    return(jaccard_n)
  })
  return(results)
}
```

```{r}
load("take_mat.RData")

nrep <- dim(take_mat)[2]

n_props <- c(0.005,0.01,0.05,0.1,0.2)
stamp_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat[,i]
  current_counts <- sample1_counts[to_take,]
  current_coords <- sample1_coords[to_take,]
  n_sizes <- floor(n_props * sum(take_mat[,i]))
  stamp_out <- stamp_wrapper_complete(X = current_counts,
                                      coords = as.matrix(current_coords))
  jc_out_stamp <- jaccard_lowdim(low_dims = list(stamp_out[[1]]),
                                 coords_list = list(stamp_out[[2]]),
                                 n_vals = n_sizes)
  stamp_nsim_mat <- stamp_nsim_mat %>% rbind(jc_out_stamp[[1]][1,])
}

# Embryo 2
load("take_mat2.RData")

n_props <- c(0.005,0.01,0.05,0.1,0.2)
stamp_nsim_mat2 <- matrix(NA, nrow = 0, ncol = length(n_props))

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat2[,i]
  current_counts <- sample2_counts[to_take,]
  current_coords <- sample2_coords[to_take,]
  n_sizes <- floor(n_props * sum(take_mat2[,i]))
  stamp_out <- stamp_wrapper_complete(X = current_counts,
                                      coords = as.matrix(current_coords))
  jc_out_stamp <- jaccard_lowdim(low_dims = list(stamp_out[[1]]),
                                 coords_list = list(stamp_out[[2]]),
                                 n_vals = n_sizes)
  stamp_nsim_mat2 <- stamp_nsim_mat2 %>% rbind(jc_out_stamp[[1]][1,])
}

# Embryo 3
load("take_mat3.RData")

n_props <- c(0.005,0.01,0.05,0.1,0.2)
stamp_nsim_mat3 <- matrix(NA, nrow = 0, ncol = length(n_props))

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat3[,i]
  current_counts <- sample3_counts[to_take,]
  current_coords <- sample3_coords[to_take,]
  n_sizes <- floor(n_props * sum(take_mat3[,i]))
  stamp_out <- stamp_wrapper_complete(X = current_counts,
                                      coords = as.matrix(current_coords))
  jc_out_stamp <- jaccard_lowdim(low_dims = list(stamp_out[[1]]),
                                 coords_list = list(stamp_out[[2]]),
                                 n_vals = n_sizes)
  stamp_nsim_mat3 <- stamp_nsim_mat3 %>% rbind(jc_out_stamp[[1]][1,])
}

#save(stamp_nsim_mat, stamp_nsim_mat2, stamp_nsim_mat3, file = "stamp_nsim_mats.RData")
```


Others

```{r}
n_props <- c(0.005,0.01,0.05,0.1,0.2)

# Embryo 1
wsir_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
precast_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
drsc_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
spca_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))

load("take_mat.RData")
nrep <- dim(take_mat)[2]
for (i in 1:nrep) {
  load(paste0("benchmark_embeddings/lowdim_out_r",i,".RData"))
  load(paste0("benchmark_embeddings/spca_out_r",i,".RData"))
  
  precast_retained <- rownames(sample1_coords) %in% rownames(lowdim_out$PRECAST)
  precast_coords <- sample1_coords[precast_retained,]
  coords_list <- list(sample1_coords[take_mat[,i],], 
                      precast_coords, 
                      sample1_coords[take_mat[,i],],
                      spca_out$coords)
  lowdim_out$SPCA = spca_out$projected
  
  n_sizes <- floor(n_props * nrow(sample1_coords[take_mat[,i],]))
  
  jc_out <- jaccard_lowdim(low_dims = lowdim_out,
                           coords_list = coords_list,
                           n_vals = n_sizes)
  wsir_nsim_mat <- wsir_nsim_mat %>% rbind(jc_out[[1]][1,])
  precast_nsim_mat <- precast_nsim_mat %>% rbind(jc_out[[2]][1,])
  drsc_nsim_mat <- drsc_nsim_mat %>% rbind(jc_out[[3]][1,])
  spca_nsim_mat <- spca_nsim_mat %>% rbind(jc_out[[4]][1,])
}

# Embryo 2
wsir_nsim_mat2 <- matrix(NA, nrow = 0, ncol = length(n_props))
precast_nsim_mat2 <- matrix(NA, nrow = 0, ncol = length(n_props))
drsc_nsim_mat2 <- matrix(NA, nrow = 0, ncol = length(n_props))
spca_nsim_mat2 <- matrix(NA, nrow = 0, ncol = length(n_props))

load("take_mat2.RData")
nrep <- dim(take_mat2)[2]
for (i in 1:nrep) {
  load(paste0("benchmark_embeddings2/lowdim_out_r",i,".RData"))
  load(paste0("benchmark_embeddings2/spca_out_r",i,".RData"))
  
  precast_retained <- rownames(sample2_coords) %in% rownames(lowdim_out$PRECAST)
  precast_coords <- sample2_coords[precast_retained,]
  coords_list <- list(sample2_coords[take_mat2[,i],], 
                      precast_coords, 
                      sample2_coords[take_mat2[,i],],
                      spca_out$coords)
  lowdim_out$SPCA = spca_out$projected
  
  n_sizes <- floor(n_props * nrow(sample2_coords[take_mat2[,i],]))
  
  jc_out <- jaccard_lowdim(low_dims = lowdim_out,
                           coords_list = coords_list,
                           n_vals = n_sizes)
  wsir_nsim_mat2 <- wsir_nsim_mat2 %>% rbind(jc_out[[1]][1,])
  precast_nsim_mat2 <- precast_nsim_mat2 %>% rbind(jc_out[[2]][1,])
  drsc_nsim_mat2 <- drsc_nsim_mat2 %>% rbind(jc_out[[3]][1,])
  spca_nsim_mat2 <- spca_nsim_mat2 %>% rbind(jc_out[[4]][1,])
}

# Embryo 3
wsir_nsim_mat3 <- matrix(NA, nrow = 0, ncol = length(n_props))
precast_nsim_mat3 <- matrix(NA, nrow = 0, ncol = length(n_props))
drsc_nsim_mat3 <- matrix(NA, nrow = 0, ncol = length(n_props))
spca_nsim_mat3 <- matrix(NA, nrow = 0, ncol = length(n_props))

load("take_mat3.RData")
nrep <- dim(take_mat3)[2]
for (i in 1:nrep) {
  load(paste0("benchmark_embeddings3/lowdim_out_r",i,".RData"))
  load(paste0("benchmark_embeddings3/spca_out_r",i,".RData"))
  
  precast_retained <- rownames(sample3_coords) %in% rownames(lowdim_out$PRECAST)
  precast_coords <- sample3_coords[precast_retained,]
  coords_list <- list(sample3_coords[take_mat3[,i],], 
                      precast_coords, 
                      sample3_coords[take_mat3[,i],],
                      spca_out$coords)
  lowdim_out$SPCA = spca_out$projected
  
  n_sizes <- floor(n_props * nrow(sample3_coords[take_mat3[,i],]))
  
  jc_out <- jaccard_lowdim(low_dims = lowdim_out,
                           coords_list = coords_list,
                           n_vals = n_sizes)
  wsir_nsim_mat3 <- wsir_nsim_mat3 %>% rbind(jc_out[[1]][1,])
  precast_nsim_mat3 <- precast_nsim_mat3 %>% rbind(jc_out[[2]][1,])
  drsc_nsim_mat3 <- drsc_nsim_mat3 %>% rbind(jc_out[[3]][1,])
  spca_nsim_mat3 <- spca_nsim_mat3 %>% rbind(jc_out[[4]][1,])
}

#load("benchmark_embeddings/stamp_nsim_mats.RData")

nsim_e1_df <- data.frame(means = c(colMeans(wsir_nsim_mat),
                                   colMeans(precast_nsim_mat),
                                   colMeans(drsc_nsim_mat),
                                   colMeans(spca_nsim_mat),
                                   colMeans(stamp_nsim_mat)),
                         sds = c(colSds(wsir_nsim_mat),
                                 colSds(precast_nsim_mat),
                                 colSds(drsc_nsim_mat),
                                 colSds(spca_nsim_mat),
                                 colSds(stamp_nsim_mat)),
                         method = vec_rep_each(c("wSIR",
                                                 "PRECAST",
                                                 "DR.SC",
                                                 "SpatialPCA",
                                                 "STAMP"), length(n_props)),
                         n_prop = rep(n_props, 5))
nsim_e2_df <- data.frame(means = c(colMeans(wsir_nsim_mat2),
                                   colMeans(precast_nsim_mat2),
                                   colMeans(drsc_nsim_mat2),
                                   colMeans(spca_nsim_mat2),
                                   colMeans(stamp_nsim_mat2)),
                         sds = c(colSds(wsir_nsim_mat2),
                                 colSds(precast_nsim_mat2),
                                 colSds(drsc_nsim_mat2),
                                 colSds(spca_nsim_mat2),
                                 colSds(stamp_nsim_mat2)),
                         method = vec_rep_each(c("wSIR",
                                                 "PRECAST",
                                                 "DR.SC",
                                                 "SpatialPCA",
                                                 "STAMP"), length(n_props)),
                         n_prop = rep(n_props, 5))
nsim_e3_df <- data.frame(means = c(colMeans(wsir_nsim_mat3),
                                   colMeans(precast_nsim_mat3),
                                   colMeans(drsc_nsim_mat3),
                                   colMeans(spca_nsim_mat3),
                                   colMeans(stamp_nsim_mat3)),
                         sds = c(colSds(wsir_nsim_mat3),
                                 colSds(precast_nsim_mat3),
                                 colSds(drsc_nsim_mat3),
                                 colSds(spca_nsim_mat3),
                                 colSds(stamp_nsim_mat3)),
                         method = vec_rep_each(c("wSIR",
                                                 "PRECAST",
                                                 "DR.SC",
                                                 "SpatialPCA",
                                                 "STAMP"), length(n_props)),
                         n_prop = rep(n_props, 5))
nsim_df <- rbind(nsim_e1_df,nsim_e2_df,nsim_e3_df)
nsim_df$sample <- vec_rep_each(c("Embryo 1","Embryo 2","Embryo 3"),25)
save(nsim_df, file = "nsim_df.RData")
```

## Xenium breast cancer dataset

Note: the dataset is to large to include in this github. To access the data, please refer to the data availability section in the manuscript.

Create matrices of which points to take at each iteration

```{r}
nrep <- 10
p_to_take <- 0.1

take_mat_xb1 <- matrix(NA, 
                       nrow = nrow(xb1_coords), 
                       ncol = nrep) # column i is which rows to take in iteration i

for (i in 1:nrep) { # fill matrix of which rows to take each repeat
  to_take <- sample(x = c(TRUE, FALSE), 
                    size = nrow(xb1_coords), 
                    replace = TRUE, 
                    prob = c(p_to_take,(1-p_to_take)))
  take_mat_xb1[,i] <- to_take
}
save(take_mat_xb1, file = "take_mat_xb1.RData")
```

```{r}
nrep <- 10
p_to_take <- 0.1
rownames(xb2_coords) <- rownames(xb2_exprs)

take_mat_xb2 <- matrix(NA, 
                       nrow = nrow(xb2_coords), 
                       ncol = nrep) # column i is which rows to take in iteration i

for (i in 1:nrep) { # fill matrix of which rows to take each repeat
  to_take <- sample(x = c(TRUE, FALSE), 
                    size = nrow(xb2_coords), 
                    replace = TRUE, 
                    prob = c(p_to_take,(1-p_to_take)))
  take_mat_xb2[,i] <- to_take
}
save(take_mat_xb2, file = "take_mat_xb2.RData")
```

### Sample 1

STAMP

```{r}
load("take_mat_xb1.RData")
xb1_coords <- readRDS(file = "xb1_coords.Rds") %>% as.data.frame()
xb1_counts <- readRDS(file = "xb1_counts.Rds")

take1 <- take_mat_xb1[,1]
counts1 <- xb1_counts[take1,]
coords1 <- xb1_coords[take1,]
start <- Sys.time()
stamp_out <- stamp_wrapper_complete(X = counts1,
                                    coords = as.matrix(coords1))
save(stamp_out, file = paste0("stamp_xb1/stamp_out_r",1,".RData"))
end <- Sys.time()
end-start

for (i in 2:10) {
  print(paste0("Current iteration: ", i))
  take <- take_mat_xb1[,i]
  counts_sampled <- xb1_counts[take,]
  coords_sampled <- xb1_coords[take,]
  stamp_out <- stamp_wrapper_complete(X = counts_sampled,
                                      coords = as.matrix(coords_sampled))
  save(stamp_out, file = paste0("stamp_xb1/stamp_out_r",i,".RData"))
}
```


BCDC

```{r}
nrep <- 10

bcdcs_wsir <- rep(0,nrep)
bcdcs_precast <- rep(0,nrep)
bcdcs_drsc <- rep(0,nrep)
bcdcs_stamp <- rep(0,nrep)
load("take_mat_xb1.RData")
rownames(xb1_coords) <- rownames(xb1_exprs)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat_xb1[,i]
  #current_counts <- xb1_coords[to_take,]
  current_coords <- xb1_coords[to_take,]
  current_ldo <- paste0("benchmark_xb1/lowdim_out_r",i,".RData")
  load(current_ldo)
  current_stamp_out <- paste0("benchmark_xb1/stamp_out_r",i,".RData")
  load(current_stamp_out)
  bcdc_out <- metric_bcdc(low_dims = lowdim_out, coords = current_coords)
  bcdcs_wsir[i] <- bcdc_out$wSIR
  bcdcs_precast[i] <- bcdc_out$PRECAST
  bcdcs_drsc[i] <- bcdc_out$DRSC
  bcdcs_stamp[i] <- energy::bcdcor(stamp_out[[1]], stamp_out[[2]])
}

means <- c(mean(bcdcs_wsir),
                               mean(bcdcs_precast),
                               mean(bcdcs_drsc),
                               mean(bcdcs_stamp))
sds <- c(sd(bcdcs_wsir),
                               sd(bcdcs_precast),
                               sd(bcdcs_drsc),
                               sd(bcdcs_stamp))

bcdc_xb1_df <- data.frame(method = c("wSIR",
                                 "PRECAST",
                                 "DR.SC",
                                 "STAMP"),
                      mean = means,
                      sd = sds
                      )

save(bcdc_xb1_df, file = "bcdc_xb1_df.RData")
```

neighbourhood similarity

```{r}
n_props <- c(0.005,0.01,0.05,0.1,0.2)

# Embryo 1
wsir_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
precast_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
drsc_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
stamp_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))

load("take_mat_xb1.RData")
nrep <- dim(take_mat_xb1)[2]
for (i in 1:nrep) {
  print(paste0("Current iteration: ",i))
  load(paste0("benchmark_xb1/lowdim_out_r",i,".RData"))
  load(paste0("benchmark_xb1/stamp_out_r",i,".RData"))
  
  precast_retained <- rownames(xb1_coords) %in% rownames(lowdim_out$PRECAST)
  precast_coords <- xb1_coords[precast_retained,]
  coords_list <- list(xb1_coords[take_mat_xb1[,i],], 
                      precast_coords, 
                      xb1_coords[take_mat_xb1[,i],],
                      stamp_out$coords)
  lowdim_out$STAMP = stamp_out$embedding
  
  n_sizes <- floor(n_props * nrow(xb1_coords[take_mat_xb1[,i],]))
  
  jc_out <- jaccard_lowdim(low_dims = lowdim_out,
                           coords_list = coords_list,
                           n_vals = n_sizes)
  wsir_nsim_mat <- wsir_nsim_mat %>% rbind(jc_out[[1]][1,])
  precast_nsim_mat <- precast_nsim_mat %>% rbind(jc_out[[2]][1,])
  drsc_nsim_mat <- drsc_nsim_mat %>% rbind(jc_out[[3]][1,])
  stamp_nsim_mat <- stamp_nsim_mat %>% rbind(jc_out[[4]][1,])
}
```

```{r}
nsim_xb1_df <- data.frame(means = c(colMeans(wsir_nsim_mat),
                                    colMeans(precast_nsim_mat),
                                    colMeans(drsc_nsim_mat),
                                    colMeans(stamp_nsim_mat)),
                         sds = c(colSds(wsir_nsim_mat),
                                 colSds(precast_nsim_mat),
                                 colSds(drsc_nsim_mat),
                                 colSds(stamp_nsim_mat)),
                         method = vec_rep_each(c("wSIR",
                                                 "PRECAST",
                                                 "DR.SC",
                                                 "STAMP"), length(n_props)),
                         n_prop = rep(n_props, 4))
save(nsim_xb1_df, file = "nsim_xb1_df.RData")
```

### Sample 2

STAMP

```{r}
load("take_mat_xb2.RData")
xb2_coords <- readRDS(file = "xb2_coords.Rds") %>% as.data.frame()
xb2_counts <- readRDS(file = "xb2_counts.Rds")

for (i in 1:10) {
  print(paste0("Current iteration: ", i))
  take <- take_mat_xb2[,i]
  counts_sampled <- xb2_counts[take,]
  coords_sampled <- xb2_coords[take,]
  stamp_out <- stamp_wrapper_complete(X = counts_sampled,
                                      coords = as.matrix(coords_sampled))
  save(stamp_out, file = paste0("stamp_xb2/stamp_out_r",i,".RData"))
}
```

BCDC

```{r}
nrep <- 10

bcdcs_wsir <- rep(0,nrep)
bcdcs_precast <- rep(0,nrep)
bcdcs_drsc <- rep(0,nrep)
bcdcs_stamp <- rep(0,nrep)
load("take_mat_xb2.RData")
rownames(xb2_coords) <- rownames(xb2_exprs)

for (i in 1:nrep) {
  print(paste0("Current iteration: ", i))
  to_take <- take_mat_xb2[,i]
  #current_counts <- xb1_coords[to_take,]
  current_coords <- xb2_coords[to_take,]
  current_ldo <- paste0("benchmark_xb2/lowdim_out_r",i,".RData")
  load(current_ldo)
  current_stamp_out <- paste0("benchmark_xb2/stamp_out_r",i,".RData")
  load(current_stamp_out)
  bcdc_out <- metric_bcdc(low_dims = lowdim_out, coords = current_coords)
  bcdcs_wsir[i] <- bcdc_out$wSIR
  bcdcs_precast[i] <- bcdc_out$PRECAST
  bcdcs_drsc[i] <- bcdc_out$DRSC
  bcdcs_stamp[i] <- energy::bcdcor(stamp_out[[1]], stamp_out[[2]])
}

means <- c(mean(bcdcs_wsir),
                               mean(bcdcs_precast),
                               mean(bcdcs_drsc),
                               mean(bcdcs_stamp))
sds <- c(sd(bcdcs_wsir),
                               sd(bcdcs_precast),
                               sd(bcdcs_drsc),
                               sd(bcdcs_stamp))

bcdc_xb2_df <- data.frame(method = c("wSIR",
                                 "PRECAST",
                                 "DR.SC",
                                 "STAMP"),
                      mean = means,
                      sd = sds
                      )

save(bcdc_xb2_df, file = "bcdc_xb2_df.RData")
```

Neighbourhood similarity

```{r}
n_props <- c(0.005,0.01,0.05,0.1,0.2)

# Embryo 1
wsir_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
precast_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
drsc_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))
stamp_nsim_mat <- matrix(NA, nrow = 0, ncol = length(n_props))

load("take_mat_xb2.RData")
nrep <- dim(take_mat_xb2)[2]
for (i in 1:nrep) {
  print(paste0("Current iteration: ",i))
  load(paste0("benchmark_xb2/lowdim_out_r",i,".RData"))
  load(paste0("benchmark_xb2/stamp_out_r",i,".RData"))
  
  precast_retained <- rownames(xb2_coords) %in% rownames(lowdim_out$PRECAST)
  precast_coords <- xb2_coords[precast_retained,]
  coords_list <- list(xb2_coords[take_mat_xb2[,i],], 
                      precast_coords, 
                      xb2_coords[take_mat_xb2[,i],],
                      stamp_out$coords)
  lowdim_out$STAMP = stamp_out$embedding
  
  n_sizes <- floor(n_props * nrow(xb2_coords[take_mat_xb2[,i],]))
  
  jc_out <- jaccard_lowdim(low_dims = lowdim_out,
                           coords_list = coords_list,
                           n_vals = n_sizes)
  wsir_nsim_mat <- wsir_nsim_mat %>% rbind(jc_out[[1]][1,])
  precast_nsim_mat <- precast_nsim_mat %>% rbind(jc_out[[2]][1,])
  drsc_nsim_mat <- drsc_nsim_mat %>% rbind(jc_out[[3]][1,])
  stamp_nsim_mat <- stamp_nsim_mat %>% rbind(jc_out[[4]][1,])
}

nsim_xb2_df <- data.frame(means = c(colMeans(wsir_nsim_mat),
                                    colMeans(precast_nsim_mat),
                                    colMeans(drsc_nsim_mat),
                                    colMeans(stamp_nsim_mat)),
                         sds = c(colSds(wsir_nsim_mat),
                                 colSds(precast_nsim_mat),
                                 colSds(drsc_nsim_mat),
                                 colSds(stamp_nsim_mat)),
                         method = vec_rep_each(c("wSIR",
                                                 "PRECAST",
                                                 "DR.SC",
                                                 "STAMP"), length(n_props)),
                         n_prop = rep(n_props, 4))
save(nsim_xb2_df, file = "nsim_xb2_df.RData")
```


