---
title: "analysis_main_fig2"
author: '500455892'
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(wSIR)
library(tidyverse)
```


Analysis for main fig 2: donut simulation, parameter robustness, benchmarking wSIR vs spatial methods

# Donut simulation

# Parameter robustness

```{r}
load("MouseData.Rdata")
coords1 <- coords_list[[1]]
coords3 <- coords_list[[3]]
exprs1 <- exprs_logcount_list[[1]]
exprs3 <- exprs_logcount_list[[3]]
alpha_vals <- c(0,4,8,15,30,50)
slice_vals <- c(5,10,15,20,30,50)

# Define parameters
alpha_vals <- c(0, 4, 8, 15, 30, 50)
slice_vals <- c(5, 10, 15, 20, 30, 50)
nrep <- 5

# Precompute lengths
n_alpha <- length(alpha_vals)
n_slices <- length(slice_vals)

# Preallocate the bc_dcs vector
total_iterations <- n_alpha * n_slices * nrep
bc_dcs <- numeric(total_iterations)

# Parallelized loop using mclapply
results <- lapply(seq_along(alpha_vals), function(a_idx) {
  alpha <- alpha_vals[a_idx]
  cat(sprintf("Current alpha: %d\n", alpha))
  
  sapply(seq_along(slice_vals), function(s_idx) {
    slices <- slice_vals[s_idx]
    cat(sprintf("  Current slices: %d\n", slices))
    
    replicate(nrep, {
      # Random sampling
      rsample1 <- sample(c(TRUE, FALSE), replace = TRUE, size = nrow(coords1), prob = c(0.8, 0.2))
      rsample3 <- sample(c(TRUE, FALSE), replace = TRUE, size = nrow(coords3), prob = c(0.8, 0.2))
      
      # Subsetting
      exprs_train <- exprs3[rsample3, ]
      coords_train <- coords3[rsample3, ]
      exprs_test <- exprs1[rsample1, ]
      coords_test <- coords1[rsample1, ]
      
      # Create wSIR object and calculate bc_dcs
      wsir_obj <- wSIR:::wSIRSpecifiedParams(X = exprs_train,
                                             coords = coords_train,
                                             alpha = alpha,
                                             slices = slices,
                                             varThreshold = 0.95)
      projected <- projectWSIR(wsir = wsir_obj, newdata = exprs_test)
      bc_distCor(projected, coords_test)
    })
  })
})

# Flatten the results into the bc_dcs vector

```

```{r}
# each column is one alpha, and each row is one slice
bc_dcs_avg <- sapply(results, colMeans)
bc_dc_params_train3test1_df <- data.frame(bc_distcor = c(bc_dcs_avg),
                                          alpha = factor(rep(alpha_vals, each = 6)),
                                          slices = factor(rep(slice_vals, 6)))

ggplot(bc_dc_params_train3test1_df, aes(x = alpha, y = slices, fill = bc_distcor)) +
  geom_tile() +
  scale_fill_viridis_c()

saveRDS(bc_dc_params_train3test1_df, file = "embryo_results_bc_dc_params_train3test1_df.Rds")
```

Xenium breast cancer

First need to preprocess the data

```{r setup, include=FALSE}
library(reticulate)
use_python("/Library/Frameworks/Python.framework/Versions/3.12/bin/python3")
source_python("graphPCA_wrap.py")
knitr::opts_chunk$set(echo = TRUE)
source("comparison_functions2.R")
```

Note the xenium data files are too large for github. To run this code, please see the data availability section of the paper and download the data from there. 

```{r, eval=F}
# Step 1: Load the .h5ad file
spe <- anndata::read_h5ad("../../data/xenium_rep1.h5ad")

# Step 2: Add per-cell QC metrics
# Assuming raw counts are stored in "X"
spe$obs$total <- rowSums(spe$X)
spe$obs$detected <- rowSums(spe$X > 0)

# Step 3: Filter cells based on QC metrics
filtered_cells <- spe$obs$total >= 10 & spe$obs$detected >= 5
spe <- spe[filtered_cells, ]

# Step 4: Examine cluster composition and subset to "Stromal" cells
cluster_counts <- table(spe$obs$Cluster) # Cluster composition
sorted_cluster_counts <- sort(cluster_counts) # Sorted cluster composition
stromal_cells <- spe$obs$Cluster == "Stromal" # Logical vector for "Stromal" cells
spe <- spe[stromal_cells, ]

counts_raw <- spe$X # Save the raw counts data for graphPCA
spe2 <- AnnData2SCE(spe)
spe2 <- logNormCounts(spe2, exprs_values = "X")

# Step 6: Save raw expression data and spatial coordinates
exprs_raw = t(assay(spe2, "logcounts"))
coords_raw = as.data.frame(reducedDim(spe2, "spatial"))
save(counts_raw, exprs_raw, coords_raw, file = "xenium_data.Rdata")
```

Begin analysis

```{r}
load("xenium_data.Rdata")

# Define parameters
alpha_vals <- c(0, 4, 8, 15, 30, 50)
slice_vals <- c(5, 10, 15, 20, 30, 50)
nrep <- 5

# Precompute lengths
n_alpha <- length(alpha_vals)
n_slices <- length(slice_vals)

# Preallocate the bc_dcs vector
total_iterations <- n_alpha * n_slices * nrep
bc_dcs <- numeric(total_iterations)

# Parallelized loop using mclapply
results <- lapply(seq_along(alpha_vals), function(a_idx) {
  alpha <- alpha_vals[a_idx]
  cat(sprintf("Current alpha: %d\n", alpha))
  
  sapply(seq_along(slice_vals), function(s_idx) {
    slices <- slice_vals[s_idx]
    cat(sprintf("  Current slices: %d\n", slices))
    
    replicate(nrep, {
      # Random sampling
      rsample1 <- sample(c(TRUE, FALSE), replace = TRUE, size = nrow(coords_raw), prob = c(0.8, 0.2))
      
      # Subsetting
      exprs_train <- exprs_raw[rsample1, ]
      coords_train <- coords_raw[rsample1, ]
      exprs_test <- exprs_raw[!rsample1, ]
      coords_test <- coords_raw[!rsample1, ]
      
      # Create wSIR object and calculate bc_dcs
      wsir_obj <- wSIR:::wSIRSpecifiedParams(X = exprs_train,
                                             coords = coords_train,
                                             alpha = alpha,
                                             slices = slices,
                                             varThreshold = 0.95)
      projected <- wSIR:::projectWSIR(wsir = wsir_obj, newdata = exprs_test)
      bc_distCor(projected, coords_test)
    })
  })
})

# Flatten the results into the bc_dcs vector

```

```{r}
# each column is one alpha, and each row is one slice
bc_dcs_avg <- sapply(results, colMeans)
bc_dc_params_train3test1_df <- data.frame(bc_distcor = c(bc_dcs_avg),
                                          alpha = factor(rep(alpha_vals, each = 6)),
                                          slices = factor(rep(slice_vals, 6)))

ggplot(bc_dc_params_df, aes(x = alpha, y = slices, fill = bc_distcor)) +
  geom_tile() +
  scale_fill_viridis_c()

saveRDS(bc_dc_params_df, file = "xenium_results_bc_dc_params.Rds")
```

# Benchmarking


